Calcineurin inhibition enhances Caenorhabditis elegans lifespan by defecation defects-mediated calorie restriction and nuclear hormone signaling

Version 1

Abstract

Abstract

Calcineurin is a highly conserved calcium/calmodulin-dependent serine/threonine protein phosphatase with diverse functions. Inhibition of calcineurin is known to enhance Caenorhabditis elegans lifespan via multiple signaling pathways. Aiming to study the role of calcineurin in regulating innate immunity, we discover that calcineurin is required for the rhythmic defecation motor program (DMP) in C. elegans. Calcineurin inhibition leads to defects in the DMP, resulting in intestinal bloating, rapid colonization of the gut by bacteria, and increased susceptibility to bacterial infection. We demonstrate that intestinal bloating by calcineurin inhibition mimics calorie restriction that results in enhanced lifespan. The TFEB ortholog, HLH-30, is required for calcineurin inhibition-mediated lifespan enhancement by triggering lipolysis. Finally, we show that the nuclear hormone receptor, NHR-8, is upregulated by calcineurin inhibition and is required for increased lifespan. Our studies uncover a role for calcineurin in the C. elegans DMP and provide a new mechanism for calcineurin inhibition-mediated longevity extension.

Introduction

Interventions that enhance lifespan also impart resistance to multiple stresses (Johnson et al., 2001). Indeed, the positive correlation between improved stress resistance and enhanced lifespan has been exploited to identify long-lived mutants (Castro et al., 2004; Denzel et al., 2014; Johnson et al., 2001; Muñoz and Riddle, 2003; Wang et al., 2004). Among stress responses, innate immunity appears to be a crucial factor for enhanced lifespan (Campos et al., 2021; Fabian et al., 2021; Soo et al., 2023; Xia et al., 2019). However, the correlation between innate immunity and lifespan is not always positive. Interventions that alter lifespan may not modulate innate immunity, and vice-versa (Labed et al., 2018; Naim et al., 2021; Otarigho and Aballay, 2020; Sun et al., 2011). Some signaling pathways also establish a tradeoff between innate immunity and lifespan. Mutants that have improved immunity but reduced lifespan have been identified (Amrit et al., 2019; Otarigho and Aballay, 2021; Ren and Ambros, 2015). Conversely, mutants with enhanced lifespans but declined immune responses have also been discovered (Kawli et al., 2010). Moreover, the genetic pathways for lifespan and immunity could be uncoupled in mutants exhibiting enhanced lifespan and improved immune responses (Alper et al., 2010; Guerrero et al., 2021). Therefore, the relationship between lifespan and innate immunity appears to be complex and remains to be fully understood.

Calcineurin, a conserved protein from yeast to humans, is a calcium/calmodulin-dependent serine/threonine protein phosphatase that is involved in diverse cellular processes and signal transduction pathways (Chen et al., 2022; Hogan et al., 2003; Schulz and Yutzey, 2004; Ulengin-Talkish and Cyert, 2023). Dephosphorylation of substrate proteins by calcineurin affects several cellular pathways, including transcriptional signaling programs (Ulengin-Talkish and Cyert, 2023). Calcineurin regulates the activity of the transcription factors of the nuclear factor of activated T cells (NFAT) family (Hogan et al., 2003). Dephosphorylation of NFATs by calcineurin triggers their nuclear localization and activates immune responses in vertebrates (Herbst et al., 2015; Hogan et al., 2003; Vandewalle et al., 2014). In the nematode Caenorhabditis elegans, calcineurin regulates thermotaxis, body size, fertility, and lifespan (Bandyopadhyay et al., 2002; Dong et al., 2007; Kuhara et al., 2002; Lee et al., 2013). Knockdown of the catalytic subunit of calcineurin, tax-6, is known to enhance C. elegans lifespan via multiple pathways, including autophagy and CREB-regulated transcriptional coactivators (CRTCs) (Dong et al., 2007; Dwivedi et al., 2009; Mair et al., 2011; Tao et al., 2013). However, the role of calcineurin in regulating C. elegans response to pathogen infections has not been studied. Because C. elegans lacks the NFAT transcription factors (Song et al., 2013), it will be intriguing to study how calcineurin inhibition impacts C. elegans innate immunity. These studies could also shed some light on the complex interplay between lifespan and immunity.

In this study, we examined the effect of calcineurin inhibition on C. elegans innate immunity. Surprisingly, we found that the knockdown of tax-6 enhanced the susceptibility of C. elegans to bacterial infection, despite enhancing lifespan. We discovered that tax-6 is required for the rhythmic defecation motor program (DMP). The knockdown of tax-6 resulted in intestinal bloating due to defects in the DMP which enhanced susceptibility to bacterial infection by increasing gut colonization by bacteria. Intestinal bloating resulted in calorie restriction-like phenotypes, including reduced lipid levels, and led to increased lifespan. We discovered that the TFEB ortholog, HLH-30, is required for calcineurin inhibition-mediated lifespan extension by triggering lipolysis. Finally, we found that the nuclear hormone receptor, NHR-8, is upregulated by calcineurin inhibition and is required for increased lifespan. Our studies uncover a new mechanism for calcineurin inhibition-mediated longevity extension.

Results

Calcineurin knockdown enhances C. elegans susceptibility to Pseudomonas aeruginosa infection

To understand the role of calcineurin in C. elegans innate immune response, we studied the survival of a hypomorphic allele of tax-6, tax-6(p675), on the pathogenic bacterium P. aeruginosa PA14. Surprisingly, the tax-6(p675) animals had a drastically reduced survival on P. aeruginosa compared to wild-type N2 animals (Figure 1A). Similarly, the knockdown of tax-6 by RNA interference (RNAi) enhanced the susceptibility of animals to P. aeruginosa compared to that of control animals (Figure 1B). The effects of RNAi were specific to tax-6, as tax-6(p675) animals did not have a further increase in susceptibility to P. aeruginosa upon tax-6 RNAi (Figure 1C). As reported earlier (Dong et al., 2007; Dwivedi et al., 2009; Mair et al., 2011; Tao et al., 2013), we confirmed that tax-6 knockdown led to an increased lifespan on E. coli (Figure 1D). An earlier study demonstrated that calcineurin regulates cAMP response element-binding protein (CREB) and CRTCs to regulate lifespan in C. elegans (Mair et al., 2011). The knockdown of the CREB homolog-1 (crh-1) and crtc-1 enhanced lifespan similar to the tax-6 knockdown (Mair et al., 2011). We asked whether the knockdown of crh-1 and crtc-1 impacted C. elegans survival on P. aeruginosa comparable to the tax-6 knockdown. Interestingly, the knockdown of crh-1 and crtc-1 did not affect the survival on P. aeruginosa as adversely as the knockdown of tax-6 did (Figure S1). These results demonstrated that despite having an increased lifespan, tax-6 knockdown animals have drastically enhanced susceptibility to P. aeruginosa infection by a mechanism likely independent of crh-1 and crtc-1.

Calcineurin knockdown enhances C. elegans susceptibility to Pseudomonas aeruginosa

(A) Representative survival plots of N2 and tax-6(p675) animals on P. aeruginosa PA14 at 25°C. The animals were grown on E. coli OP50 at 20°C until 1-day-old adults before transferring to P. aeruginosa PA14 at 25°C. p<0.001

(B) Representative survival plots of N2 animals on P. aeruginosa PA14 at 25°C after treatment with the empty vector (EV) control and tax-6 RNAi. p<0.001.

(C) Representative survival plots of tax-6(p675) animals on P. aeruginosa PA14 at 25°C after treatment with the EV control and tax-6 RNAi. n.s., nonsignificant.

(D) Representative survival plots of N2 animals grown on bacteria for RNAi against tax-6 along with the EV control at 20°C. Day 0 represents young adults. p<0.001.

(E) Representative survival plots of fer-1(b232) animals on P. aeruginosa PA14 at 25°C after treatment with the EV control and tax-6 RNAi at 25°C. p<0.001.

(F) Representative survival plots of fer-1(b232) animals grown on bacteria for RNAi against tax-6 along with the EV control at 25°C. The animals were developed at 25°C. Day 0 represents young adults. p<0.001.

For all panels, n = 3 biological replicates; animals per condition per replicate > 90.

We observed enhanced matricidal hatching in tax-6 knockdown animals on P. aeruginosa. Therefore, we asked whether the enhanced susceptibility of tax-6 knockdown animals to P. aeruginosa was because of enhanced matricidal hatching. To this end, we studied the effects of tax-6 knockdown in fer-1(b232) temperature-sensitive mutants. When grown at 25°C, fer-1(b232) animals have unfertilized oocytes (Argon and Ward, 1980), and therefore, these animals will lack matricidal hatching. As shown in Figure 1E, fer-1(b232) animals with unfertilized oocytes also had reduced survival on P. aeruginosa upon tax-6 knockdown, indicating that the increased susceptibility upon tax-6 knockdown is not because of enhanced matricidal hatching. The fer-1(b232) animals also showed enhanced lifespan on E. coli upon knockdown of tax-6 (Figure 1F). These results suggested that the knockdown of tax-6 led to increased lifespan and enhanced susceptibility to pathogen infection.

Next, we asked whether the enhanced susceptibility to bacterial infection upon tax-6 knockdown was mediated by one or more of the established C. elegans innate immunity pathways, which include a MAP kinase pathway mediated by NSY-1/SEK-1/PMK-1 (Kim et al., 2002), the MLK-1/MEK-1/KGB-1 c-Jun kinase pathway (Kim et al., 2004), the TGF-β/DBL-1 pathway (Mallo et al., 2002), and the bZIP transcription factor ZIP-2 pathway (Estes et al., 2010). In mutants of all different immune pathways, the knockdown of tax-6 further enhanced susceptibility to P. aeruginosa (Figure S2A-D). The enhanced susceptibility of tax-6 knockdown animals to P. aeruginosa was also independent of the FOXO transcription factor DAF-16 (Figure S2E). Thus, the enhanced susceptibility to P. aeruginosa upon tax-6 knockdown appears to be independent of established C. elegans innate immunity pathways.

Calcineurin is required for the C. elegans defecation motor program (DMP)

To understand why tax-6 knockdown animals had enhanced susceptibility to P. aeruginosa infection, we studied the colonization of the intestine of tax-6 RNAi animals by P. aeruginosa. Knockdown of tax-6 led to enhanced intestinal colonization by P. aeruginosa compared to the control animals (Figure 2A, B). The tax-6(p675) animals also had increased intestinal colonization by P. aeruginosa (Figure 2C, D). Animals with bloated intestinal lumens exhibit enhanced bacterial colonization of the gut, an improved lifespan, and an increased susceptibility to pathogens (Kumar et al., 2019; Singh and Aballay, 2019a). Because tax-6 knockdown animals displayed all of these phenotypes, we asked whether tax-6 knockdown led to the bloating of the intestinal lumen. Indeed, we found that tax-6 knockdown resulted in bloated intestinal lumens (Figure 2E, F).

Calcineurin is required for the C. elegans defecation motor program (DMP)

(A) Representative fluorescence (top) and the corresponding brightfield (bottom) images of N2 animals incubated on P. aeruginosa-GFP for 6 hours at 25°C after growth on the empty vector (EV) control and tax-6 RNAi bacteria.

(B) Colony-forming units (CFU) per animal of N2 worms incubated on P. aeruginosa-GFP for 6 hours at 25°C after growth on the EV control and tax-6 RNAi bacteria. **p < 0.01 via the t-test (n = 3 biological replicates).

(C) Representative fluorescence (top) and the corresponding brightfield (bottom) images of N2 and tax-6(p675) animals incubated on P. aeruginosa-GFP for 6 hours at 25°C after growth on E. coli OP50 at 20°C.

(D) CFU per animal of N2 and tax-6(p675) worms incubated on P. aeruginosa-GFP for 6 hours at 25°C after growth on E. coli OP50 at 20°C. *p < 0.05 via the t-test (n = 3 biological replicates).

(E) Representative photomicrographs of N2 animals grown on the EV control and tax-6 RNAi bacteria at 20°C until 1-day-old adults. Arrows point to the border of the intestinal lumen.

(F) Quantification of the diameter of the intestinal lumen of N2 animals grown on the EV control and tax-6 RNAi bacteria at 20°C until 1-day-old adults. ***p<0.001 via the t-test (n = 21 worms each).

(G) The number of expulsion events observed in 20 minutes in 1-day-old adult N2 animals grown on the EV control and tax-6 RNAi bacteria at 20°C. ***p<0.001 via the t-test (n = 6 worms each).

(H) Percent of 1-day-old adult worms having irregular DMP. ***p<0.001 via the t-test (n = 4 biological replicates).

Intestinal bloating could be an outcome of either defects in the pharyngeal pumping (Kumar et al., 2019) or defects in the defecation motor program (DMP) (Singh and Aballay, 2019a). Knockdown of tax-6 did not affect the pharyngeal pumping rate (Figure S3A). We studied whether calcineurin was required for the DMP. The C. elegans DMP is a highly coordinated rhythmic behavior required for regular expulsion of the intestinal lumen contents with a frequency of about one cycle per minute. The DMP involves a series of muscle contractions involving posterior body muscle contraction, anterior body muscle contraction, and expulsion muscle contraction to release the lumen content (Thomas, 1990). We counted the number of expulsion events for 20 minutes per animal and found that tax-6 knockdown animals had drastically reduced expulsion events (Figure 2G), leading to irregular DMP. The DMP-defect phenotype had a low penetrance in the tax-6(p675) animals. The tax-6(p675) animals exhibited both regular and irregular DMPs (Figure S3B). About 36% of the tax-6(p675) animals had irregular and slowed DMP, while the rest had a regular DMP (Figure 2H), suggesting that tax-6(p675) is a weak allele. Taken together, these results showed that calcineurin is required for the DMP, and its inhibition enhances susceptibility to P. aeruginosa by impeding the DMP.

Calcineurin inhibition enhances lifespan via DMP defects-mediated calorie restriction

Defects in C. elegans DMP result in reduced absorption of nutrients leading to diminished intestinal fat (Sheng et al., 2015). We asked whether tax-6 knockdown resulted in reduced fat levels. To this end, we carried out oil-red-O (ORO) staining of tax-6 RNAi animals. Knockdown of tax-6 resulted in drastically reduced fat levels (Figure 3A, B). Next, we tested whether the reduced fat levels in tax-6 knockdown animals resulted in calorie restriction-like phenotypes. Calorie restriction is known to upregulate the expression of the FoxA transcription factor PHA-4 (Panowski et al., 2007). Knockdown of tax-6 also resulted in the upregulation of pha-4 expression levels (Figure 3C). Because calorie restriction enhances lifespan (Kaeberlein et al., 2006; Lakowski and Hekimi, 1998; Panowski et al., 2007), we asked whether calcineurin inhibition resulted in increased lifespan via calorie restriction. The knockdown of tax-6 in the genetic model of calorie restriction, eat-2(ad465) (Lakowski and Hekimi, 1998), did not increase the lifespan (Figure 3D). These results suggested that calcineurin inhibition enhances lifespan via calorie restriction.

Calcineurin knockdown enhances lifespan via DMP defects-mediated calorie restriction

(A) Representative photomicrographs of oil-red-O (ORO) stained 1-day-old adult N2 animals grown on the empty vector (EV) control and tax-6 RNAi bacteria at 20°C. Scale bar = 200 µm.

(B) Quantification of ORO intensity per animal of 1-day-old adult N2 animals grown on the EV control and tax-6 RNAi bacteria at 20°C. ***p<0.001 via the t-test (n = 10 worms each).

(C) Quantitative reverse transcription-PCR (qRT-PCR) for pha-4 mRNA levels in N2 animals grown on the EV control and tax-6 RNAi bacteria at 20°C until 1-day-old adults. ***p < 0.001 (n = 3 biological replicates).

(D) Representative survival plots of eat-2(ad465) animals grown on the bacteria for RNAi against tax-6 along with the EV control at 20°C. Day 0 represents young adults. n.s., nonsignificant (n = 3 biological replicates; animals per condition per replicate > 90).

(E-I) Representative survival plots of aak-2(ok524) (E), raga-1(ok386) (F), rsks-1(ok1255) (G), daf-16(mu86) (H), and nhr-49(nr2041) (I) animals grown on the bacteria for RNAi against tax-6 along with the EV control at 20°C. Day 0 represents young adults. p<0.001 for all the plots (n = 3 biological replicates; animals per condition per replicate > 80).

The mechanisms of lifespan enhancement by dietary or calorie restriction are complex and depend on several different downstream pathways (Chamoli et al., 2014; Greer and Brunet, 2009; Hansen et al., 2008). Further, different pathways are operational under different dietary regimens (Greer and Brunet, 2009). Therefore, we tested the role of different pathways linked with calorie restriction-mediated lifespan enhancement in increasing lifespan upon calcineurin inhibition. We studied the interactions of the AMP-activated kinase, aak-2 (Greer and Brunet, 2009), the mTOR pathway, raga-1 (Robida-Stubbs et al., 2012; Zhang et al., 2019), ribosomal S6 kinase, rsks-1 (Chen et al., 2009; Selman et al., 2009; Zhang et al., 2019), the FOXO transcription factor, daf-16 (Greer and Brunet, 2009), and the nuclear hormone receptor, nhr-49 (Chamoli et al., 2014), with calcineurin inhibition in regulating lifespan. Mutants defective in aak-2 appeared to have only a partial increase in lifespan upon calcineurin inhibition (Figure 3E). On the other hand, loss-of-function mutants of raga-1, rsks-1, daf-16, and nhr-49 exhibited enhanced lifespan upon calcineurin inhibition (Figure 3F-I).

Calcineurin knockdown enhances lifespan via HLH-30 and NHR-8

To further explore longevity pathways regulated by calorie restriction that might enhance lifespan downstream of calcineurin inhibition, we studied the role of TFEB ortholog HLH-30. HLH-30 is required for starvation response and triggers lipid depletion under nutrient-deprived conditions (O’Rourke and Ruvkun, 2013). Knockdown of tax-6 did not enhance the lifespan of hlh-30(tm1978) animals (Figure 4A), indicating that HLH-30 is required for increased lifespan downstream of calcineurin inhibition. We studied whether HLH-30 affected fat depletion upon the knockdown of tax-6. While the ORO levels declined significantly in N2 animals upon knockdown of tax-6 (Figure 3A, B), the ORO levels did not change in hlh-30(tm1978) animals upon tax-6 knockdown (Figure 4B, C). Knockdown of tax-6 resulted in defects in the DMP and intestinal bloating in hlh-30(tm1978) animals (Figure S4A-C). Therefore, the lack of fat depletion in hlh-30(tm1978) animals upon tax-6 RNAi was not because of the absence of DMP defects. These results suggested that HLH-30-mediated lipid depletion enhanced lifespan upon calcineurin inhibition.

Calcineurin inhibition enhances lifespan via HLH-30 and NHR-8

(A) Representative survival plots of hlh-30(tm1978) animals grown on the bacteria for RNAi against tax-6 along with the empty vector (EV) control at 20°C. Day 0 represents young adults. n.s., nonsignificant (n = 3 biological replicates; animals per condition per replicate > 90).

(B) Representative photomicrographs of oil-red-O (ORO) stained 1-day-old adult hlh-30(tm1978) animals grown on the EV control and tax-6 RNAi bacteria at 20°C. Scale bar = 200 µm.

(C) Quantification of ORO intensity per animal of 1-day-old adult hlh-30(tm1978) animals grown on the EV control and tax-6 RNAi bacteria at 20°C. n.s., nonsignificant via the t-test (n = 10 worms each).

(D) Representative survival plots of nhr-8(ok186) animals grown on the bacteria for RNAi against tax-6 along with the EV control at 20°C. Day 0 represents young adults. n.s., nonsignificant (n = 3 biological replicates; animals per condition per replicate > 90).

(E) Representative survival plots of daf-9(rh50) animals grown on the bacteria for RNAi against tax-6 along with the EV control at 20°C. Day 0 represents young adults. p<0.001 (n = 3 biological replicates; animals per condition per replicate > 80).

(F) Representative photomicrographs of ORO stained 1-day-old adult nhr-8(ok186) animals grown on the EV control and tax-6 RNAi bacteria at 20°C. Scale bar = 200 µm.

(G) Quantification of ORO intensity per animal of 1-day-old adult nhr-8(ok186) animals grown on the EV control and tax-6 RNAi bacteria at 20°C. ***p<0.001 via the t-test (n = 10 worms each).

(H) Quantitative reverse transcription-PCR (qRT-PCR) for nhr-8 mRNA levels in N2 animals grown on the EV control, tax-6, flr-1, nhx-2, and pbo-1 RNAi bacteria at 20°C until 1-day-old adults. ***p < 0.001 via the t-test (n = 3 biological replicates).

(I) Model depicting the mechanism of increased lifespan and enhanced susceptibility to pathogen upon inhibition of calcineurin via the defects in the defecation motor program (DMP).

Steroid hormone signaling mediated by NHR-8 is known to regulate calorie restriction-mediated enhancement in lifespan (Thondamal et al., 2014). We tested whether NHR-8 was required for lifespan increment by calcineurin inhibition. Indeed, the knockdown of tax-6 did not enhance the lifespan of nhr-8(ok186) animals (Figure 4D). While the ligands of NHR-8 remain poorly defined (Magner et al., 2013), the steroid hormones produced by the cytochrome P450 enzyme DAF-9 are known to enhance lifespan via NHR-8 under calorie restriction (Thondamal et al., 2014). Therefore, we tested whether DAF-9 mediated the enhanced lifespan downstream of calcineurin inhibition. We found that tax-6 knockdown enhanced the lifespan of daf-9(rh50) animals (Figure 4E), indicating that DAF-9-produced steroid hormones were not required for lifespan increment by calcineurin inhibition. We asked whether NHR-8 affected the fat depletion of tax-6 knockdown animals. The nhr-8(ok186) animals had drastically reduced fat levels upon the knockdown of tax-6 (Figure 4F, G). These results indicated that NHR-8 worked downstream of or in parallel with fat depletion.

Because calorie restriction can also enhance the expression of genes that increase lifespan, such as pha-4 (Panowski et al., 2007), we tested whether tax-6 knockdown modulated the mRNA levels of nhr-8. Indeed, we observed that tax-6 knockdown increased nhr-8 mRNA levels significantly (Figure 4H). To test whether tax-6 knockdown resulted in the upregulation of nhr-8 because of intestinal bloating or independent of intestinal bloating, we studied nhr-8 mRNA levels in other gene knockdown animals that had bloated intestinal lumens. We knocked down genes flr-1, nhx-2, and pbo-1 as their inhibition is known to lead to defects in the DMP, resulting in intestinal bloating and enhanced lifespan (Singh and Aballay, 2019a). Knockdown of flr-1, nhx-2, and pbo-1 also resulted in increased nhr-8 mRNA levels (Figure 4H), indicating that intestinal bloating resulted in nhr-8 upregulation. Therefore, calcineurin inhibition likely enhanced lifespan via NHR-8 by upregulating its expression levels. Taken together, these results suggested that intestinal bloating caused by calcineurin inhibition enhances lifespan by lipolysis via HLH-30 and upregulating expression of NHR-8 (Figure 4I).

Discussion

We discovered that calcineurin is required for the C. elegans DMP. Calcineurin activity is regulated by intracellular calcium levels. Increased amounts of calcium ions activate the phosphatase activity of calcineurin via the binding of the calcium-sensing protein calmodulin to calcineurin (Klee et al., 1998). The rhythmic DMP cycle in C. elegans is known to be regulated by rhythmic calcium waves (Santo et al., 1999; Teramoto and Iwasaki, 2006). The calcium waves are regulated by the endoplasmic reticulum calcium channel, ITR-1, and mutations in the itr-1 gene affect defecation by preventing cytoplasmic calcium release (Santo et al., 1999). It is likely that calcium waves regulate DMP via calcineurin activities. Indeed, calcineurin is expressed in the enteric muscles that are required for contractions for DMP (Lee et al., 2005). It is interesting to note that tax-6 gain-of-function mutants are also known to have DMP defects (Lee et al., 2005). Therefore, optimum calcineurin activity appears to be crucial for maintaining a rhythmic DMP.

Calcineurin inhibition has been shown to extend C. elegans lifespan via multiple mechanisms (Dong et al., 2007; Dwivedi et al., 2009; Mair et al., 2011; Tao et al., 2013). Activation of autophagy has been shown to be one of the mechanisms of extended lifespan upon calcineurin inhibition (Dwivedi et al., 2009). We showed that defects in the DMP caused by calcineurin inhibition reduce lipid levels and mimic calorie restriction. Calorie restriction is known to induce autophagy (Morselli et al., 2010), and autophagy is required for calorie restriction-mediated lifespan extension (Hansen et al., 2008; Jia and Levine, 2007). Therefore, it is likely that reduced lipid levels because of intestinal bloating result in the activation of autophagy upon calcineurin inhibition. Importantly, the TFEB ortholog, HLH-30, is required for lipolysis and autophagy under starvation conditions (Lapierre et al., 2013; O’Rourke and Ruvkun, 2013). HLH-30 is also required for lifespan enhancement via autophagy and multiple longevity pathways (Lapierre et al., 2013; O’Rourke and Ruvkun, 2013). We observed that HLH-30 is also required for the increase in lifespan upon calcineurin inhibition. Because autophagy may regulate lifespan via multiple longevity pathways (Hansen et al., 2008; Lapierre et al., 2013), this could explain why calcineurin inhibition enhances lifespan via multiple mechanisms.

Dietary restriction-mediated lifespan extensions are complex and context-dependent (Greer and Brunet, 2009). Multiple longevity pathways have been identified under different paradigms of dietary restriction (Chamoli et al., 2020, 2014; Chen et al., 2009; Greer and Brunet, 2009; Lapierre et al., 2013; Matai et al., 2019; Selman et al., 2009; Thondamal et al., 2014). We found that tax-6 knockdown enhanced lifespan independent of multiple dietary restriction pathways, including raga-1, rsks-1, daf-16, and nhr-49. An earlier study identified that tax-6 knockdown enhanced lifespan via NHR-49 (Burkewitz et al., 2015). We currently do not understand the reasons for this discrepancy. We identified that tax-6 knockdown resulted in the upregulation of nhr-8 mRNA levels, and nhr-8 was required for the increased lifespan in tax-6 knockdown animals. NHR-8 has been shown to mediate calorie restriction-dependent lifespan via the regulation of xenobiotic responses (Chamoli et al., 2014; Verma et al., 2018). Therefore, it is likely that NHR-8 increases lifespan upon calcineurin inhibition by activating xenobiotic response pathways.

Recent studies in different organisms have shown that gut bloating has profound effects on food-seeking behaviors, immunity, and lifespan (Duvall et al., 2019; Filipowicz et al., 2021; Kumar et al., 2019; Min et al., 2021; Singh and Aballay, 2019a, 2019b, 2019c). Several C. elegans mutants with defects in the DMP and bloated intestinal lumens are known to have dampened nutrient absorption, leading to reduced lipid deposition in the gut, mimicking calorie restriction (Sheng et al., 2015). It is possible that some of the effects of intestinal bloating on the host physiology are mediated by calorie restriction. Indeed, the neuropeptide Y receptors, which control a diverse set of behaviors, including appetite, are activated by gut bloating (Singh and Aballay, 2019c). Calorie restriction is known to induce neuropeptide Y, which might trigger feeding behaviors (Aveleira et al., 2015; de Rijke et al., 2005; Ferreira-Marques et al., 2016). The complete characterization of the physiological changes downstream of gut bloating may provide broad insights into the effects of gut physiology on behaviors, immunity, and lifespan.

Materials and Methods

Bacterial strains

The following bacterial strains were used: Escherichia coli OP50, E. coli HT115(DE3), Pseudomonas aeruginosa PA14, and P. aeruginosa PA14 expressing green fluorescent protein (P. aeruginosa PA14-GFP). The cultures for these bacteria were grown in Luria-Bertani (LB) broth at 37°C.

C. elegans strains and growth conditions

C. elegans hermaphrodites were maintained on nematode growth medium (NGM) plates seeded with E. coli OP50 at 20°C unless otherwise indicated. Bristol N2 hermaphrodites were used as the wild-type control unless otherwise indicated. The following strains were used in the study: PR675 tax-6(p675), HH142 fer-1(b232), DA465 eat-2(ad465), CF1038 daf-16(mu86), STE68 nhr-49(nr2041), RB754 aak-2(ok524), AE501 nhr-8(ok186), VC222 raga-1(ok386), RG1228 daf-9(rh50), RB1206 rsks-1(ok1255), KU25 pmk-1(km25), KU21 kgb-1(km21), NU3 dbl-1(nk3), JIN1375 hlh-30(tm1978), and VC3056 zip-2(ok3730). Some of the strains were obtained from the Caenorhabditis Genetics Center (University of Minnesota, Minneapolis, MN). The fer-1(b232) hermaphrodites were maintained on E. coli OP50 at 15°C and were grown at 25°C prior to P. aeruginosa killing assays and longevity assays to obtain animals with unfertilized oocytes.

RNA interference (RNAi)

RNAi was used to generate loss-of-function phenotypes by feeding nematodes E. coli strain HT115(DE3) expressing double-stranded RNA homologous to a target gene. RNAi was carried out as described previously (Gokul and Singh, 2022; Ravi et al., 2023). Briefly, E. coli with the appropriate vectors were grown in LB broth containing ampicillin (100 μg/mL) at 37°C overnight and plated onto NGM plates containing 100 μg/mL ampicillin and 3 mM isopropyl β-D-thiogalactoside (IPTG) (RNAi plates). RNAi-expressing bacteria were allowed to grow overnight at 37°C. Gravid adults were transferred to RNAi-expressing bacterial lawns and allowed to lay eggs for 2 hours. The gravid adults were removed, and the eggs were allowed to develop at 20°C to 1-day-old adults for subsequent assays. The RNAi clones were from the Ahringer RNAi library and were verified by sequencing.

C. elegans longevity assays

Lifespan assays were performed on RNAi plates containing E. coli HT115(DE3) with the empty vector control and tax-6 RNAi clone in the presence of 50 µg/mL of 5-fluorodeoxyuridine (FUdR). Animals were synchronized on RNAi plates without FUdR and incubated at 20°C. At the late L4 larval stage, the animals were transferred onto the corresponding RNAi plates containing 50 µg/mL of FUdR and incubated at 20°C. For fer-1(b232) lifespan assays, the animals were synchronized on RNAi plates without FUdR and incubated at 25°C. The fer-1(b232) lifespan assays were carried out at 25°C without FUdR. Animals were scored every other day as live, dead, or gone. Animals that failed to display touch-provoked movement were scored as dead. Animals that crawled off the plates were censored. Experimental groups contained more than 80 animals per condition per replicate. Young adult animals were considered as day 0 for the lifespan analysis. Three independent experiments were performed.

C. elegans killing assays on P. aeruginosa PA14

Bacterial cultures were prepared by inoculating individual bacterial colonies of P. aeruginosa into 2 mL of LB and growing them for 10-12 hours on a shaker at 37°C. Bacterial lawns were prepared by spreading 20 µL of the culture on the entire surface of 3.5-cm-diameter modified NGM agar plates (0.35% instead of 0.25% peptone). The plates were incubated at 37°C for 12-16 hours and then cooled to room temperature for at least 30 minutes before seeding with synchronized 1-day-old adult animals. The killing assays were performed at 25°C, and live animals were transferred daily to fresh plates. Animals were scored at times indicated and were considered dead when they failed to respond to touch.

P. aeruginosa-GFP colonization assay

Bacterial cultures were prepared by inoculating individual bacterial colonies of P. aeruginosa-GFP into 2 mL of LB and growing them for 10-12 hours on a shaker at 37°C. Bacterial lawns were prepared by spreading 20 µL of the culture on the entire surface of 3.5-cm-diameter modified NGM agar plates (0.35% instead of 0.25% peptone) containing 50 μg/mL of kanamycin. The plates were incubated at 37°C for 12 hours and then cooled to room temperature for at least 30 minutes before seeding with 1-day-old adult gravid adults. The assays were performed at 25°C. After 6 hours of incubation, the animals were transferred from P. aeruginosa-GFP plates to fresh E. coli OP50 plates and visualized within 5 minutes under a fluorescence microscope.

Quantification of intestinal bacterial loads

The intestinal bacterial loads were quantified by measuring colony-forming units (CFU) as described earlier (Singh and Aballay, 2019b, 2019a) with appropriate modifications. Briefly, P. aeruginosa-GFP lawns were prepared as described above. After 6 hours of exposure, the animals were transferred from P. aeruginosa-GFP plates to the center of fresh E. coli OP50 plates thrice for 10 minutes each to eliminate bacteria stuck to their body. Afterward, ten animals/condition were transferred into 50 μL of PBS containing 0.01% triton X-100 and ground using glass beads. Serial dilutions of the lysates (101, 102, 103, 104) were seeded onto LB plates containing 50 μg/mL of kanamycin to select for P. aeruginosa-GFP cells and grown overnight at 37°C. Single colonies were counted the next day and represented as the number of bacterial cells or CFU per animal. At least three independent experiments were performed for each condition.

Fluorescence imaging

Fluorescence imaging was carried out as described previously (Gokul and Singh, 2022; Ravi et al., 2023). Briefly, the animals were anesthetized using an M9 salt solution containing 50 mM sodium azide and mounted onto 2% agarose pads. The animals were then visualized using a Nikon SMZ-1000 fluorescence stereomicroscope. The fluorescence intensity was quantified using Image J software.

Quantification of intestinal lumen bloating

The 1-day-old adult animals grown on the empty vector control and tax-6 RNAi were anesthetized using an M9 salt solution containing 50 mM sodium azide, mounted onto 2% agarose pads, and imaged. The diameter of the intestinal lumen was measured using the Zeiss Zen Pro 2011 software. At least ten animals were used for each condition.

Measurement of defecation motor program (DMP) rate

Wild-type N2 and tax-6(p675) animals were synchronized and grown at 20°C on E. coli OP50 for 4 and 5 days, respectively, before measuring the DMP rate. For RNAi experiments, N2 worms were synchronized on EV and tax-6 RNAi plates and grown for 4 days at 20°C before measuring the DMP rate. The DMP cycle length was scored by assessing the time between expulsions (which are preceded by posterior and anterior body wall muscle contraction and the contraction of enteric muscles in a normal regular pattern) (Thomas, 1990). The number of expulsion events in 20 minutes was measured for each worm. The DMP rate was recorded for 5-6 worms/condition.

Pharyngeal pumping assay

For the pharyngeal pumping assay, 1-day-old adult animals grown on the empty vector control and tax-6 RNAi were used. The number of contractions of the terminal bulb of the pharynx was counted for 30 seconds per worm. The pumping rates for 20 worms were recorded for each condition.

Lipid staining

The oil-red-O (ORO) staining was carried out as described earlier (Lynn et al., 2015) with appropriate modification. Briefly, wild-type N2, hlh-30(tm1978), and nhr-8(ok186) animals were synchronized on empty vector control and tax-6 RNAi bacteria and grown at 20°C for 4 days. About 300-400 synchronized gravid adult worms from experimental plates were washed three times with 1X PBS plus 0.01% triton X-100 (PBST). To permeabilize the cuticle, 600 µL of 40% isopropanol was added to 100 µL of worm pellet and was rocked for three minutes. The animals were spun down at 500 rpm for 30 seconds, and then 600 µL of supernatant was removed. Subsequently, 600 µL of ORO working stock was added to the worm pellet to stain the worms and incubated at room temperature for 1 hour on a shaker. ORO working stock was freshly prepared by diluting the stock 0.5% ORO in isopropanol to 60% in water and rocked at room temperature for 2 hours, followed by the removal of debris with a 0.22 µm filter. Afterward, worm samples were pelleted, 600 µL of supernatant was removed, 600 µL of PBST was added, and the samples were rocked for 30 minutes at room temperature. After that, the worms were mounted on a 2% agarose pad and imaged. At least ten worms/condition were imaged. Three independent biological replicates were performed. The ORO intensity was quantified using Image J software.

RNA isolation and quantitative reverse transcription-PCR (qRT-PCR)

Animals were synchronized by egg laying. Approximately 40 N2 gravid adult animals were transferred to 9-cm RNAi plates seeded with E. coli HT115 expressing the appropriate vectors and allowed to lay eggs for 4-5 hours. The gravid adults were then removed, and the eggs were allowed to develop at 20°C for 96 hours. Subsequently, the animals were collected, washed with M9 buffer, and frozen in TRIzol reagent (Life Technologies, Carlsbad, CA). Total RNA was extracted using the RNeasy Plus Universal Kit (Qiagen, Netherlands). Residual genomic DNA was removed using TURBO DNase (Life Technologies, Carlsbad, CA). A total of 6 μg of total RNA was reverse-transcribed with random primers using the High-Capacity cDNA Reverse Transcription Kit (Applied Biosystems, Foster City, CA).

qRT-PCR was conducted using the Applied Biosystems One-Step Real-time PCR protocol using SYBR Green fluorescence (Applied Biosystems) on an Applied Biosystems 7900HT real-time PCR machine in 96-well-plate format. Twenty-five-microliter reactions were analyzed as outlined by the manufacturer (Applied Biosystems). The relative fold-changes of the transcripts were calculated using the comparative CT(2-ΔΔCT) method and normalized to pan-actin (act-1, -3, -4) as described earlier (Singh and Aballay, 2019a, 2017). The cycle thresholds of the amplification were determined using StepOnePlus software (Applied Biosystems). All samples were run in triplicate. The primer sequences are shown in Table S1.

Quantification and statistical analysis

The statistical analysis was performed with Prism 8 (GraphPad). All error bars represent mean±standard deviation (SD). The two-sample t-test was used when needed, and the data were judged to be statistically significant when p < 0.05. In the figures, asterisks (*) denote statistical significance as follows: *, p<0.05, **, p<0.001, ***, p<0.0001, as compared with the appropriate controls. The Kaplan-Meier method was used to calculate the survival fractions, and statistical significance between survival curves was determined using the log-rank test. All experiments were performed in triplicate.

eLife Assessment

This useful study has the potential to reveal insights into how calcineurin influences C. elegans lifespan through its role in controlling the defecation motor program. Currently, the evidence in support of the conclusions is still incomplete, largely due to concerns about partial gene inactivation by RNAi. The inclusion of experiments using a tax-6 null allele would mitigate these concerns.

Reviewer #1 (Public Review):

In this paper, the authors show that disruption of calcineurin, which is encoded by tax-6 in C. elegans, results in increased susceptibility to P. aeruginosa, but extends lifespan. In exploring the mechanisms involved, the authors show that disruption of tax-6 decreases the rate of defecation leading to intestinal accumulation of bacteria and distension of the intestinal lumen. The authors further show that the lifespan extension is dependent on hlh-30, which may be involved in breaking down lipids following deficits in defecation, and nhr-8, whose levels are increased by deficits in defecation. The authors propose a model in which disruption of the defecation motor program is responsible for the effect of calcineurin on pathogen susceptibility and lifespan, but do not exclude the possibility that calcineurin affects these phenotypes independently of defecation.

Reviewer #2 (Public Review):

The manuscript titled "Calcineurin Inhibition Enhances Caenorhabditis elegans Lifespan by Defecation Defects-Mediated Calorie Restriction and Nuclear Hormone Signaling" by Priyanka Das, Alejandro Aballay, and Jogender Singh reveals that inhibiting calcineurin, a conserved protein phosphatase, in C. elegans affects the defecation motor program (DMP), leading to intestinal bloating and increased susceptibility to bacterial infection. This intestinal bloating mimics calorie restriction, ultimately resulting in an enhanced lifespan. The research identifies the involvement of HLH-30 and NHR-8 proteins in this lifespan enhancement, providing new insights into the role of calcineurin in C. elegans DMP and mechanisms for longevity.

The authors present novel findings on the role of calcineurin in regulating the defecation motor program in C. elegans and how its inhibition can lead to lifespan enhancement. The evidence provided is solid with multiple experiments supporting the main claims.

Strengths:

The manuscript's strength lies in the authors' use of genetic and biochemical techniques to investigate the role of calcineurin in regulating the DMP, innate immunity, and lifespan in C. elegans. Moreover, the authors' findings provide a new mechanism for calcineurin inhibition-mediated longevity extension, which could have significant implications for understanding the molecular basis of aging and developing interventions to promote healthy aging.

1. The study uncovers a new role for calcineurin in the regulation of C. elegans DMP and a potential novel pathway for enhancing lifespan via calorie restriction involving calcineurin, HLH-30, and NHR-8 in C. elegans.

2. Multiple signaling pathways involved in lifespan enhancement were investigated with fairly strong experimental evidence supporting their claims.

Weaknesses:

The manuscript's weaknesses include the lack of mechanistic details regarding how calcineurin inhibition leads to defects in the DMP and induces calorie restriction-like effects on lifespan.

The exact site of calcineurin action, i.e., whether in the intestine or enteric muscles (Lee et al., 2005), and the possible molecular mechanisms linking calcineurin inhibition, DMP defects, and lifespan were not adequately explored. Although characterization of the full mechanism is probably beyond the scope of this paper, given the relative simplicity and advantages of using C. elegans as a model organism for this study, some degree of rigor is expected with additional straightforward control experiments as listed below:

The authors state that tax-6 knockdown animals had drastically reduced expulsion events (Figure 2G), leading to irregular DMP (Lines 144-145), but did not describe the nature of DMP irregularity. For example, did the reduced expulsion events still occur with regular intervals but longer cycle lengths? Or was the rhythmicity completely abolished? The former would suggest the intestine clock is still intact, and the latter would indicate that calcineurin is required for the clock to function. Therefore, ethograms of DMP in both wild-type and tax-6 mutant animals are warranted to be included in the manuscript. Along the same line, besides the cycle length, the three separable motor steps (aBoc, pBoc, EMC) are easily measurable, with each step indicating where the program goes wrong, hence the site of action, which is precisely the beauty of studying C. elegans DMP. Unfortunately, the authors did not use this opportunity to characterize the exact behavior phenotypes of the tax-6 mutant to guide future investigations. Furthermore, it is interesting that about 64% of tax-6 (p675) animals had normal DMP. The authors attributed this to p675 being a weak allele. It would be informative to further examine tax-6 RNAi as in other experiments or to make a tax-6 null mutant with CRISPR. In addition, in one of the cited papers (Lee et al., 2005), the exact calcineurin loss-of-function strain tax-6(p675) was shown to have normal defecation, including normal EMC, while the gain-of-function mutant of calcineurin tax-6(jh107) had abnormal EMC steps. It wasn't clear from Lee et al., 2005, if the reported "normal defecation" was only referring to the expulsion step or also included the cycle length. Nevertheless, this potential contradiction and calcineurin gain-of-function mutant is highly relevant to the current study and should be further explored as a follow-up to previously reported results. For some of the key experiments, such as tax-6's effects on susceptibility to PA14, DMP, intestinal bloating, and lifespan, additional controls, as the norm of C. elegans studies, including second allele and rescue experiments, would strengthen the authors' claims and conclusions.

The second weakness of this manuscript is the data presentation for all survival rate curves. The authors stated that three independent experiments or biological replicates were performed for each group but only showed one "representative" curve for each plot. Without seeing all individual datasets or the averaged data with error bars, there is no way to evaluate the variability and consistency of the survival rate reported in this study.

Overall, the authors' claims and conclusions are justified by their data, but further experiments are needed to confirm their findings and establish the detailed mechanisms underlying the observed effects of calcineurin inhibition on the DMP, calorie restriction, and lifespan in C. elegans.

Author response:

The following is the authors’ response to the original reviews.

We would like to thank the reviewers for their interest in our studies. In response to their comments, we have conducted additional experiments and made the necessary revisions to the manuscript. The new studies included to address the reviewers’ comments are shown in Figure 1B, 1F, Figure 2—figure supplement 1, Figure 3, Figure 3—figure supplement 1, Figure 3—figure supplement 2, Figure 3—figure supplement 3, Figure 4E, Figure 4—figure supplement 1, Figure 5, Figure 5—figure supplement 1, Figure 5—figure supplement 2D, and Figure 6. We are grateful for the critiques, which have helped us substantially improve the quality of the manuscript.

Below, we have provided a point-by-point response to the reviewers’ comments.

Public Reviews:

Reviewer #1 (Public Review):

In this paper, the authors show that disruption of calcineurin, which is encoded by tax-6 in C. elegans, results in increased susceptibility to P. aeruginosa, but extends lifespan. In exploring the mechanisms involved, the authors show that disruption of tax-6 decreases the rate of defecation leading to intestinal accumulation of bacteria and distension of the intestinal lumen. The authors further show that the lifespan extension is dependent on hlh-30, which may be involved in breaking down lipids following deficits in defecation, and nhr-8, whose levels are increased by deficits in defecation. The authors propose a model in which disruption of the defecation motor program is responsible for the effect of calcineurin on pathogen susceptibility and lifespan, but do not exclude the possibility that calcineurin affects these phenotypes independently of defecation.

We thank the reviewer for providing an excellent summary of our work. We have performed additional experiments as suggested by both the reviewers and believe we have thoroughly addressed all the reviewers' concerns.

Reviewer #2 (Public Review):

The manuscript titled "Calcineurin Inhibition Enhances Caenorhabditis elegans Lifespan by Defecation Defects-Mediated Calorie Restriction and Nuclear Hormone Signaling" by Priyanka Das, Alejandro Aballay, and Jogender Singh reveals that inhibiting calcineurin, a conserved protein phosphatase, in C. elegans affects the defecation motor program (DMP), leading to intestinal bloating and increased susceptibility to bacterial infection. This intestinal bloating mimics calorie restriction, ultimately resulting in an enhanced lifespan. The research identifies the involvement of HLH-30 and NHR-8 proteins in this lifespan enhancement, providing new insights into the role of calcineurin in C. elegans DMP and mechanisms for longevity.

The authors present novel findings on the role of calcineurin in regulating the defecation motor program in C. elegans and how its inhibition can lead to lifespan enhancement. The evidence provided is solid with multiple experiments supporting the main claims.

Strengths:

The manuscript's strength lies in the authors' use of genetic and biochemical techniques to investigate the role of calcineurin in regulating the DMP, innate immunity, and lifespan in C. elegans. Moreover, the authors' findings provide a new mechanism for calcineurin inhibitionmediated longevity extension, which could have significant implications for understanding the molecular basis of aging and developing interventions to promote healthy aging.

(1) The study uncovers a new role for calcineurin in the regulation of C. elegans DMP and a potential novel pathway for enhancing lifespan via calorie restriction involving calcineurin, HLH-30, and NHR-8 in C. elegans.

(2) Multiple signaling pathways involved in lifespan enhancement were investigated with fairly strong experimental evidence supporting their claims.

We thank the reviewer for an excellent summary of our work and for highlighting the strengths of the findings.

Weaknesses:

The manuscript's weaknesses include the lack of mechanistic details regarding how calcineurin inhibition leads to defects in the DMP and induces calorie restriction-like effects on lifespan.

The exact site of calcineurin action, i.e., whether in the intestine or enteric muscles (Lee et al., 2005), and the possible molecular mechanisms linking calcineurin inhibition, DMP defects, and lifespan were not adequately explored. Although characterization of the full mechanism is probably beyond the scope of this paper, given the relative simplicity and advantages of using C. elegans as a model organism for this study, some degree of rigor is expected with additional straightforward control experiments as listed below:

The authors state that tax-6 knockdown animals had drastically reduced expulsion events (Figure 2G), leading to irregular DMP (Lines 144-145), but did not describe the nature of DMP irregularity. For example, did the reduced expulsion events still occur with regular intervals but longer cycle lengths? Or was the rhythmicity completely abolished? The former would suggest the intestine clock is still intact, and the latter would indicate that calcineurin is required for the clock to function. Therefore, ethograms of DMP in both wild-type and tax6 mutant animals are warranted to be included in the manuscript. Along the same line, besides the cycle length, the three separable motor steps (aBoc, pBoc, EMC) are easily measurable, with each step indicating where the program goes wrong, hence the site of action, which is precisely the beauty of studying C. elegans DMP. Unfortunately, the authors did not use this opportunity to characterize the exact behavior phenotypes of the tax-6 mutant to guide future investigations. Furthermore, it is interesting that about 64% of tax-6 (p675) animals had normal DMP. The authors attributed this to p675 being a weak allele. It would be informative to further examine tax-6 RNAi as in other experiments or to make a tax-6 null mutant with CRISPR. In addition, in one of the cited papers (Lee et al., 2005), the exact calcineurin loss-of-function strain tax-6(p675) was shown to have normal defecation, including normal EMC, while the gain-of-function mutant of calcineurin tax-6(jh107) had abnormal EMC steps. It wasn't clear from Lee et al., 2005, if the reported "normal defecation" was only referring to the expulsion step or also included the cycle length. Nevertheless, this potential contradiction and calcineurin gain-of-function mutant is highly relevant to the current study and should be further explored as a follow-up to previously reported results. For some of the key experiments, such as tax-6's effects on susceptibility to PA14, DMP, intestinal bloating, and lifespan, additional controls, as the norm of C. elegans studies, including second allele and rescue experiments, would strengthen the authors' claims and conclusions.

We have now included lifespan, survival on P. aeruginosa, and DMP data using an additional knockout allele, tax-6(ok2065). Additionally, we have added ethograms of DMP for both tax-6 RNAi and the tax-6(ok2065) mutant. Our observations indicate that tax-6 inhibition leads to a complete loss of DMP rhythmicity, suggesting that calcineurin is essential for maintaining the DMP clock. While characterizing the DMP, we noticed that expulsion events appeared superficial in the tax-6(ok2065) mutant, with little to no gut content released. Consequently, we examined the movement of gut content and found that both tax-6(ok2065) mutants and tax-6 knockdown animals showed significantly reduced gut content movement. The new findings on the characterization of DMP are presented in Figure 2—figure supplement 1, Figure 3, Figure 3—figure supplement 1, and Figure 3—figure supplement 2. The text in the results section reads (lines 160-176): “Next, we investigated whether the reduced number of expulsion events was due to regular intervals with longer cycle lengths or if rhythmicity was entirely disrupted upon tax-6 knockdown. To assess this, we obtained ethograms of the DMP for N2 animals grown on control and tax-6 RNAi. While animals on control RNAi displayed regular cycles of pBoc, aBoc, and EMC, the tax-6 RNAi animals exhibited disrupted rhythmicity (Figure 3A and Figure 3—figure supplement 1). Most tax-6 knockdown animals lacked the pBoc and aBoc steps and had sporadic expulsion events. Isolated pBoc events were occasionally observed, indicating a complete loss of rhythmicity in tax-6 knockdown animals. Ethograms for tax-6(ok2065) animals also showed disrupted rhythmicity (Figure 3B and Figure 3—figure supplement 2). Although the number of expulsion events appeared higher in tax-6(ok2065) animals compared to tax-6 RNAi animals (Figure 3—figure supplement 1 and 2), these expulsion events seemed superficial, releasing little to no gut content. This suggested slow movement of gut content in tax6(ok2065) animals, leading to constipation and intestinal bloating. We examined gut content movement by measuring the clearance of blue dye (erioglaucine disodium salt) from the gut. The clearance was significantly slower in tax-6(ok2065) animals compared to N2 animals (Figure 3C), indicating impaired gut content movement due to the loss of tax-6. Similarly, tax-6 knockdown animals also showed significantly slowed gut content movement (Figure 3D).”

Moreover, we have added a potential reason for the tax-6(p675) contradictory results from Lee et al., 2005 (lines 154-159): “At the 1-day-old adult stage, about 36% of tax-6(p675) animals showed irregular and slowed DMP, while the remainder had regular DMP (Figure 2H), suggesting that tax-6(p675) is a weak allele. The fraction of the animals with irregular DMP appeared to increase with age, indicating that this phenotype might be agedependent. This may also explain why tax-6(p675) animals were reported to have a normal defecation cycle in an earlier study (Lee et al., 2005).”

The second weakness of this manuscript is the data presentation for all survival rate curves. The authors stated that three independent experiments or biological replicates were performed for each group but only showed one "representative" curve for each plot. Without seeing all individual datasets or the averaged data with error bars, there is no way to evaluate the variability and consistency of the survival rate reported in this study.

We now provide all replicates data in the source data files.

Overall, the authors' claims and conclusions are justified by their data, but further experiments are needed to confirm their findings and establish the detailed mechanisms underlying the observed effects of calcineurin inhibition on the DMP, calorie restriction, and lifespan in C. elegans.

We have conducted additional experiments to elucidate the role of calcineurin in the DMP and to investigate the impact of the DMP on calorie restriction and lifespan in C. elegans, as described in the various responses to the reviewers’ comments.

Recommendations for the authors:

Our specific comments to guide the authors, should they choose to revise the manuscript:

The RNAi experiments in the eat-2 mutant background are difficult to interpret. If these animals are eating fewer bacteria, it is possible that there is also less tax-6 dsRNA being ingested and therefore less tax-6 inactivation. These experiments should be conducted with a tax-6 null allele.

We have included lifespan experiments with the eat-2(ad465);tax-6(ok2065) double mutant, along with the individual single mutant controls, as shown in Figure 4E. These results demonstrate that the eat-2 mutation does not further extend the lifespan of the tax-6(ok2065) mutant. Additionally, we confirmed that the eat-2(ad465) mutants do not exhibit defects in feeding-based RNAi (Figure 4—figure supplement 1).

While aak-2, hlh-30, and nhr-8 mutants may not have an eat phenotype, the negative tax-6 RNAi results should be confirmed with a tax-6 null mutant to obviate the consideration that these background mutations reduce RNAi efficacy.

The genes hlh-30 and nhr-8 are located very close to tax-6 on chromosome IV (https://wormbase.org//#012-34-5), which made it challenging to generate double mutants. However, we tested the RNAi sensitivity of the hlh-30(tm1978) and nhr-8(ok186) mutants and confirmed that they are not defective in RNAi (Figure 5—figure supplement 1). We also found that tax-6 RNAi disrupted the DMP in both hlh-30(tm1978) and nhr-8(ok186) mutants (Figure 5—figure supplement 2). Furthermore, our results show that hlh-30(tm1978) and nhr-8(ok186) animals have increased susceptibility to P. aeruginosa upon tax-6 knockdown (Figure 6A, B), indicating that tax-6 RNAi was effective in these mutants. Since the phenotype in the aak-2 mutant was only partially observed, we did not conduct further experiments with aak-2 mutants.

Reviewer #1 (Recommendations For The Authors):

The low penetrance of defecation cycle defects in tax-6(p675) worms brings into question the role of the defecation deficits in the phenotypes caused by the disruption of tax6. At the same time, the low penetrance provides a golden opportunity to test this. Do tax6(p675) worms with a normal defecation cycle length have extended longevity? Increased susceptibility to bacterial pathogens? Smaller body size? Distended lumen? Decreased fat accumulation? Increased pha-4 and nhr-8 expression? It would be relatively straightforward to measure defecation cycle length in individual tax-6(p675) worms, bin them into normal defecation and slow defecation groups, and then compare the above-mentioned phenotypes.

We appreciate the reviewer's interesting suggestion. However, the DMP defect phenotype in tax-6(p675) worms appears to be age-dependent, with the number of DMPdefective worms increasing as they age. Additionally, we observed that exposure to P. aeruginosa accelerates the onset of DMP defects in tax-6(p675) worms. As a result, tax6(p675) worms are not suitable for the type of experiments the reviewer suggested. Nevertheless, we believe that the additional data using the tax-6(ok2065) mutant, along with the characterization of ethograms of DMP, firmly establishes the role of calcineurin in maintaining a regular DMP in C. elegans.

Another way to dissect specific effects of calcineurin disruption from phenotypes resulting from defecation motor program deficits would be to further characterize other worms with deficits in defecation (flr-1, nhx-2, pbo-1 RNAi). It is mentioned that they have decreased lifespan. Do they also show increased susceptibility to bacterial pathogens? Do they show decreased fat? Is their lifespan dependent on HLH-30 and NHR-8?

We thank the reviewer for this important suggestion. We have now included data with flr-1, nhx-2, and pbo-1 RNAi, which shows that the knockdown of these genes also enhances susceptibility to P. aeruginosa (Figure 3—figure supplement 3G). Knockdown of these genes is already known to reduce fat levels in N2 worms, and we demonstrate that they similarly reduce fat levels in hlh-30(tm1978) and nhr-8(ok186) animals (Figure 5B, C, F, G). Additionally, we found that the increased lifespan observed upon knockdown of these genes (as well as with tax-6 knockdown) is dependent on HLH-30 and NHR-8 (Figure 5A, D).

To place "enhanced susceptibility to pathogen" within the proposed model, it would be important to examine the effect of HLH-30 and NHR-8 disruption on this phenotype. The proposed model suggests that this phenotype is independent of HLH-30 and NHR-8, but this should be tested experimentally. Similarly, it would be important to test the effect of HLH-30 and NHR-8 disruption on defecation cycle length to determine if defecation deficits are upstream or downstream of deficits in the defecation motor program

We show that the knockdown of tax-6 leads to defects in the DMP in hlh30(tm1978) and nhr-8(ok186) animals (Figure 5—figure supplement 2). Moreover, we show that hlh-30(tm1978) and nhr-8(ok186) animals have increased susceptibility to P. aeruginosa upon tax-6 knockdown (Figure 6A, B). These results are described as (lines 279-285): “Given that HLH-30 and NHR-8 are essential for lifespan extension upon calcineurin inhibition, we investigated whether these pathways also influence survival in response to P. aeruginosa infection following calcineurin knockdown. Both hlh-30(tm1978) and nhr-8(ok186) animals showed significantly reduced survival upon tax-6 RNAi (Figure 6A, B). These findings suggested that the reduced survival on P. aeruginosa following calcineurin inhibition is independent of HLH-30 and NHR-8 and is more likely due to increased gut colonization by P. aeruginosa resulting from DMP defects (Figure 6C).”

Is the lifespan of tax-6(p675) increased? This would be important to measure and include in Figure 1.

Indeed, the lifespan of tax-6(p675) mutants is increased. We have included the lifespan of tax-6(p675) and tax-6(ok2065) in Figure 1F.

In Figure 2, disruption of tax-6 appears to result in a clear decrease in body size. To what extent is the decrease in fat/worm in Figure 3 simply a result of the worms being smaller? Perhaps, a measurement of Oil-Red-O intensity PER AREA would be a more appropriate measure.

The ORO intensity values we had shown per animal were already area normalized. We have now indicated this in the Figure Legends.

There are multiple long-lived mutant strains such as clk-1 and isp-1 that have an increased defecation cycle length. To what extent do these worms exhibit phenotypes similar to tax-6 disruption? isp-1 have increased resistance to bacterial pathogens suggesting that defecation motor program deficits are not sufficient to increase susceptibility to bacterial pathogens.

We have now examined the clk-1 and isp-1 mutants and found that these mutants exhibit reduced gut colonization by P. aeruginosa compared to N2 animals. This reduction in colonization may be attributed to the slowed pharyngeal pumping rates observed in these mutants. These findings suggest that the phenotypes associated with a slow DMP versus a disrupted DMP could be significantly different. The manuscript with the new data on these mutants reads (lines 177-192): “We then explored whether the disruption of DMP rhythmicity due to tax-6 knockdown affected P. aeruginosa responses similarly to longer but regular DMP cycles. To do this, we studied P. aeruginosa colonization in clk-1(qm30) and isp1(qm150) mutants, which have regular but extended DMP cycles (Feng et al., 2001; Wong et al., 1995). Interestingly, both clk-1(qm30) and isp-1(qm150) mutants showed significantly reduced intestinal colonization by P. aeruginosa compared to N2 animals (Figure 3—figure supplement 3A-D). This reduced colonization could be attributed to their significantly decreased pharyngeal pumping rates (Wong et al., 1995; Yee et al., 2014), suggesting a lower intake of bacterial food in these mutants. While the survival of clk-1(qm30) animals on P. aeruginosa was comparable to N2 animals (Figure 3—figure supplement 3E), isp1(qm150) animals exhibited significantly improved survival (Figure 3—figure supplement 3F). Conversely, knockdown of flr-1, nhx-2, and pbo-1 in N2 animals resulted in significantly reduced survival on P. aeruginosa compared to control RNAi (Figure 3—figure supplement 3G). Knockdown of these genes causes complete disruption of DMP rhythmicity, increasing gut colonization by P. aeruginosa (Singh and Aballay, 2019a). Overall, these findings demonstrated that calcineurin is crucial for maintaining the DMP ultradian clock, and its inhibition increases susceptibility to P. aeruginosa by disrupting the DMP.”

Line 192. This statement is speculative. There is no evidence that HLH-30 is mediating lipid depletion in these worms.

We have removed this statement. We observed that the knockdown of flr-1, nhx2, and pbo-1 resulted in significant fat depletion in hlh-30(tm1978) animals (Figure 5B, C). Additionally, tax-6 knockdown also caused a small but significant reduction in fat levels in hlh-30(tm1978) animals. This contrasts with our initial submission, possibly due to the increased number of animals included in the analysis. These findings suggest that the increase in lifespan due to DMP defects requires HLH-30, likely through a mechanism independent of HLH-30’s role in fat depletion. We have updated the manuscript text and model (Figure 6C) accordingly.

In Figure S2, tax-6 RNAi appears to have a more detrimental effect in pmk-1 mutants than the other mutants. The authors should comment on this.

We have added the following sentence in the manuscript (lines 123-125): “The knockdown of tax-6 appeared to have a more pronounced effect in pmk-1(km25) mutants than in other mutants, suggesting that inhibition of tax-6 might exacerbate the adverse effects observed in pmk-1(km25) mutants.”

Reviewer #2 (Recommendations For The Authors):

Line 192-193: The statement is confusing and not accurate because HLH-30 did not enhance lifespan with or without calcineurin (Figure 4A and S4A, also in Lapierre 2023). The takeaway should be along the lines of calcineurin inhibition enhancing lifespan through HLH-30 or HLH-30 being required for lifespan enhancement via calcineurin inhibition.

We have removed this statement. We now state (lines 237-239): “Knockdown of tax-6 did not extend the lifespan of hlh-30(tm1978) animals (Figure 5A), indicating that HLH-30 is required for the increased lifespan observed with calcineurin inhibition.”

Line 261: Similar to the point above. Where is the data showing NHR-8 increases lifespan with or without calcineurin?

We have removed this sentence.

Figure 1 legend line 699: animals per condition per replicate >90, but in the Method section Line 317, it says more than 80 animals per condition per replicate. Could be more accurate.

We have now specified in the Methods section that the exact number of animals per condition is provided in the source data files. Since different lifespan curves within a given figure panel had varying numbers of animals, we have indicated the lower boundary for all curves (including the replicates). The precise number of animals for each lifespan experiment is available in the source data files.

Figures 2F and G, "tax-6" should be labeled as "tax-6 RNAi" to be consistent with other figures.

We thank the reviewer for this suggestion and have updated the label to “tax-6 RNAi”.

In summary, we would like to thank the reviewers again for providing constructive critiques. We believe we have fully addressed all the concerns of the reviewers by carrying out several new experiments and modifying the text. The manuscript has undergone substantial revision and has thereby improved significantly. We do hope that the evidence in support of the conclusions is found to be complete in the revised manuscript.

Version 2

Abstract

Abstract

Calcineurin is a highly conserved calcium/calmodulin-dependent serine/threonine protein phosphatase with diverse functions. Inhibition of calcineurin is known to enhance the lifespan of Caenorhabditis elegans through multiple signaling pathways. Aiming to study the role of calcineurin in regulating innate immunity, we discover that calcineurin is required for the rhythmic defecation motor program (DMP) in C. elegans. Calcineurin inhibition leads to defects in the DMP, resulting in intestinal bloating, rapid colonization of the gut by bacteria, and increased susceptibility to bacterial infection. We demonstrate that intestinal bloating caused by calcineurin inhibition mimics the effects of calorie restriction, resulting in enhanced lifespan. The TFEB ortholog, HLH-30, is required for lifespan extension mediated by calcineurin inhibition. Finally, we show that the nuclear hormone receptor, NHR-8, is upregulated by calcineurin inhibition and is necessary for the increased lifespan. Our studies uncover a role for calcineurin in the C. elegans DMP and provide a new mechanism for calcineurin inhibition-mediated longevity extension.

Introduction

Interventions that enhance lifespan also impart resistance to multiple stresses (Johnson et al., 2001). Indeed, the positive correlation between improved stress resistance and enhanced lifespan has been exploited to identify long-lived mutants (Castro et al., 2004; Denzel et al., 2014; Johnson et al., 2001; Muñoz and Riddle, 2003; Wang et al., 2004). Among stress responses, innate immunity appears to be a crucial factor for enhanced lifespan (Campos et al., 2021; Fabian et al., 2021; Soo et al., 2023; Xia et al., 2019). However, the correlation between innate immunity and lifespan is not always positive. Interventions that alter lifespan may not modulate innate immunity, and vice-versa (Labed et al., 2018; Naim et al., 2021; Otarigho and Aballay, 2020; Sun et al., 2011). Some signaling pathways also establish a tradeoff between innate immunity and lifespan. Mutants that have improved immunity but reduced lifespan have been identified (Amrit et al., 2019; Otarigho and Aballay, 2021; Ren and Ambros, 2015). Conversely, mutants with enhanced lifespans but declined immune responses have also been discovered (Kawli et al., 2010). Moreover, the genetic pathways for lifespan and immunity could be uncoupled in mutants exhibiting enhanced lifespan and improved immune responses (Alper et al., 2010; Guerrero et al., 2021). Therefore, the relationship between lifespan and innate immunity appears to be complex and remains to be fully understood.

Calcineurin, a conserved protein from yeast to humans, is a calcium/calmodulin-dependent serine/threonine protein phosphatase that is involved in diverse cellular processes and signal transduction pathways (Chen et al., 2022; Hogan et al., 2003; Schulz and Yutzey, 2004; Ulengin-Talkish and Cyert, 2023). Dephosphorylation of substrate proteins by calcineurin affects several cellular pathways, including transcriptional signaling programs (Ulengin-Talkish and Cyert, 2023). Calcineurin regulates the activity of the transcription factors of the nuclear factor of activated T cells (NFAT) family (Hogan et al., 2003). Dephosphorylation of NFATs by calcineurin triggers their nuclear localization and activates immune responses in vertebrates (Herbst et al., 2015; Hogan et al., 2003; Vandewalle et al., 2014). In the nematode Caenorhabditis elegans, calcineurin regulates thermotaxis, body size, fertility, and lifespan (Bandyopadhyay et al., 2002; Dong et al., 2007; Kuhara et al., 2002; Lee et al., 2013). Knockdown of the catalytic subunit of calcineurin, tax-6, is known to enhance C. elegans lifespan via multiple pathways, including autophagy and CREB-regulated transcriptional coactivators (CRTCs) (Dong et al., 2007; Dwivedi et al., 2009; Mair et al., 2011; Tao et al., 2013). However, the role of calcineurin in regulating C. elegans response to pathogen infections has not been studied. Because C. elegans lacks the NFAT transcription factors (Song et al., 2013), it will be intriguing to study how calcineurin inhibition impacts C. elegans innate immunity. These studies could also shed some light on the complex interplay between lifespan and immunity.

In this study, we examined the effect of calcineurin inhibition on C. elegans innate immunity. Surprisingly, we found that the knockdown of tax-6 enhanced the susceptibility of C. elegans to bacterial infection despite enhancing lifespan. We discovered that tax-6 is required for the rhythmic defecation motor program (DMP). The knockdown of tax-6 resulted in intestinal bloating due to defects in the DMP, which enhanced susceptibility to bacterial infection by increasing gut colonization by bacteria. Intestinal bloating resulted in calorie restriction-like phenotypes, including reduced lipid levels, and led to increased lifespan. We discovered that the TFEB ortholog, HLH-30, is required for calcineurin inhibition-mediated lifespan extension. Moreover, we found that the nuclear hormone receptor, NHR-8, is upregulated by calcineurin inhibition and is necessary for increased lifespan. Our studies uncover a new mechanism for calcineurin inhibition-mediated longevity extension.

Results

Calcineurin knockdown enhances C. elegans susceptibility to Pseudomonas aeruginosa infection

To understand the role of calcineurin in the innate immune response of C. elegans, we examined the survival of a hypomorphic allele of tax-6, tax-6(p675), when exposed to the pathogenic bacterium P. aeruginosa PA14. Surprisingly, tax-6(p675) mutants showed significantly reduced survival on P. aeruginosa compared to wild-type N2 animals (Figure 1A). Similarly, animals carrying a null allele of tax-6, tax-6(ok2065), also exhibited a significant decrease in survival on P. aeruginosa compared to N2 animals (Figure 1B). In addition, knockdown of tax-6 by RNA interference (RNAi) increased susceptibility to P. aeruginosa relative to control animals (Figure 1C). The RNAi effects were specific to tax-6, as tax-6(p675) mutants did not show further increased susceptibility to P. aeruginosa when subjected to tax-6 RNAi (Figure 1D). As reported earlier (Dong et al., 2007; Dwivedi et al., 2009; Mair et al., 2011; Tao et al., 2013), we confirmed that both tax-6 loss of function and RNAi knockdown led to an increased lifespan on E. coli (Figure 1E, F). An earlier study demonstrated that calcineurin regulates cAMP response element-binding protein (CREB) and CRTCs to regulate lifespan in C. elegans (Mair et al., 2011). The knockdown of the CREB homolog-1 (crh-1) and crtc-1 enhanced lifespan similar to the tax-6 knockdown (Mair et al., 2011). We investigated whether crh-1 and crtc-1 knockdowns also affected C. elegans survival on P. aeruginosa as severely as tax-6 knockdown. Interestingly, crh-1 and crtc-1 knockdowns did not compromise survival on P. aeruginosa as much as tax-6 knockdown did (Figure 1). These findings suggested that, despite the increased lifespan, tax-6 knockdown animals exhibit markedly enhanced susceptibility to P. aeruginosa infection through a mechanism likely independent of crh-1 and crtc-1.

Calcineurin knockdown enhances C. elegans susceptibility to P. aeruginosa

(A) Representative survival plots of N2 and tax-6(p675) animals on P. aeruginosa PA14 at 25°C. The animals were grown on E. coli OP50 at 20°C until 1-day-old adults before transferring to P. aeruginosa PA14 at 25°C. p<0.001.

(B) Representative survival plots of N2 and tax-6(ok2065) animals on P. aeruginosa PA14 at 25°C. The animals were grown on E. coli OP50 at 20°C until 1-day-old adults before transferring to P. aeruginosa PA14 at 25°C. p<0.001.

(C) Representative survival plots of N2 animals on P. aeruginosa PA14 at 25°C after treatment with the empty vector (EV) control and tax-6 RNAi. p<0.001.

(D) Representative survival plots of tax-6(p675) animals on P. aeruginosa PA14 at 25°C after treatment with the EV control and tax-6 RNAi. n.s., nonsignificant.

(E) Representative survival plots of N2 animals grown on bacteria for RNAi against tax-6 along with the EV control at 20°C. Day 0 represents young adults. p<0.001.

(F) Representative survival plots of N2, tax-6(p675), and tax-6(ok2065) animals grown on E. coli OP50 at 20°C. Day 0 represents young adults. p<0.001 for tax-6(p675) and tax-6(ok2065) compared to N2.

(G) Representative survival plots of fer-1(b232) animals on P. aeruginosa PA14 at 25°C after treatment with the EV control and tax-6 RNAi at 25°C. p<0.001.

(H) Representative survival plots of fer-1(b232) animals grown on bacteria for RNAi against tax-6 along with the EV control at 25°C. The animals were developed at 25°C. Day 0 represents young adults. p<0.001. For all panels, n = 3 biological replicates; animals per condition per replicate > 60.

We observed enhanced matricidal hatching in tax-6 knockdown animals on P. aeruginosa. Therefore, we asked whether the enhanced susceptibility of tax-6 knockdown animals to P. aeruginosa was because of enhanced matricidal hatching. To this end, we studied the effects of tax-6 knockdown in fer-1(b232) temperature-sensitive mutants. When grown at 25°C, fer-1(b232) animals have unfertilized oocytes (Argon and Ward, 1980), thus eliminating the possibility of matricidal hatching. As shown in Figure 1G, even in the absence of matricidal hatching, fer-1(b232) animals exhibited reduced survival on P. aeruginosa following tax-6 knockdown, indicating that the increased susceptibility is not due to enhanced matricidal hatching. Moreover, fer-1(b232) animals showed an extended lifespan on E. coli when tax-6 was knocked down (Figure 1H). These results suggested that tax-6 knockdown leads to both increased lifespan and enhanced susceptibility to pathogen infection.

We then explored whether the increased susceptibility to bacterial infection following tax-6 knockdown was mediated by any known C. elegans innate immunity pathways, including the MAP kinase pathway mediated by NSY-1/SEK-1/PMK-1 (Kim et al., 2002), the MLK-1/MEK-1/KGB-1 c-Jun kinase pathway (Kim et al., 2004), the TGF-β/DBL-1 pathway (Mallo et al., 2002), and the bZIP transcription factor ZIP-2 pathway (Estes et al., 2010). In mutants deficient in each of these pathways, tax-6 knockdown further increased susceptibility to P. aeruginosa (Figure 2A-D). The knockdown of tax-6 appeared to have a more pronounced effect in pmk-1(km25) mutants than in other mutants, suggesting that inhibition of tax-6 might exacerbate the adverse effects observed in pmk-1(km25) mutants. Additionally, the increased susceptibility to P. aeruginosa seen in tax-6 knockdown animals was independent of the FOXO transcription factor DAF-16 (Figure 2E). Thus, the enhanced susceptibility to P. aeruginosa resulting from tax-6 knockdown appears to operate independently of established C. elegans innate immunity pathways.

Calcineurin is required for the C. elegans defecation motor program (DMP)

To understand why tax-6 knockdown animals had enhanced susceptibility to P. aeruginosa infection, we studied the colonization of the intestine of tax-6 RNAi animals by P. aeruginosa. Knockdown of tax-6 led to enhanced intestinal colonization by P. aeruginosa compared to the control animals (Figure 2A, B). Similarly, tax-6(ok2065) and tax-6(p675) animals also showed increased intestinal colonization by P. aeruginosa compared to N2 animals (Figure 2C, D and Figure 3A, B). Animals with bloated intestinal lumens exhibit enhanced bacterial colonization of the gut, an improved lifespan, and an increased susceptibility to pathogens (Kumar et al., 2019; Singh and Aballay, 2019a). Because tax-6 knockdown animals displayed all of these phenotypes, we asked whether tax-6 knockdown led to the bloating of the intestinal lumen. Indeed, we found that tax-6 knockdown resulted in bloated intestinal lumens (Figure 2E, F).

Calcineurin is required for the C. elegans defecation motor program (DMP)

(A) Representative fluorescence (top) and the corresponding brightfield (bottom) images of N2 animals incubated on P. aeruginosa-GFP for 6 hours at 25°C after growth on the empty vector (EV) control and tax-6 RNAi bacteria. Scale bar = 200 µm.

(B) Colony-forming units (CFU) per animal of N2 worms incubated on P. aeruginosa-GFP for 6 hours at 25°C after growth on the EV control and tax-6 RNAi bacteria. **p < 0.01 via the t-test (n = 3 biological replicates).

(C) Representative fluorescence (top) and the corresponding brightfield (bottom) images of N2 and tax-6(p675) animals incubated on P. aeruginosa-GFP for 6 hours at 25°C after growth on E. coli OP50 at 20°C. Scale bar = 200 µm.

(D) CFU per animal of N2 and tax-6(p675) worms incubated on P. aeruginosa-GFP for 6 hours at 25°C after growth on E. coli OP50 at 20°C. *p < 0.05 via the t-test (n = 3 biological replicates).

(E) Representative photomicrographs of N2 animals grown on the EV control and tax-6 RNAi bacteria at 20°C until 1-day-old adults. Arrows point to the border of the intestinal lumen.

(F) Quantification of the diameter of the intestinal lumen of N2 animals grown on the EV control and tax-6 RNAi bacteria at 20°C until 1-day-old adults. ***p<0.001 via the t-test (n = 21 worms each).

(G) The number of expulsion events observed in 20 minutes in 1-day-old adult N2 animals grown on the EV control and tax-6 RNAi bacteria at 20°C. ***p<0.001 via the t-test (n = 6 worms each).

(H) Percent of 1-day-old adult worms having irregular DMP. ***p<0.001 via the t-test (n = 4 biological replicates).

Intestinal bloating could result from either defects in pharyngeal pumping (Kumar et al., 2019) or defects in the defecation motor program (DMP) (Singh and Aballay, 2019a). We found that tax-6 knockdown did not affect the pharyngeal pumping rate (Figure 4A). This prompted us to explore whether calcineurin is necessary for the DMP. The C. elegans DMP is a highly coordinated rhythmic behavior required for the regular expulsion of intestinal contents, occurring approximately once per minute. The DMP involves a series of muscle contractions, including posterior body muscle contraction (pBoc), anterior body muscle contraction (aBoc), and expulsion muscle contraction (EMC), to release intestinal contents (Thomas, 1990). When we counted expulsion events over a 20-minute period, we found that tax-6 knockdown animals had a drastically reduced number of expulsion events (Figure 2G). The DMP-defect phenotype had low penetrance in tax-6(p675) mutants, which exhibited both regular and irregular DMPs (Figure 4B). At the 1-day-old adult stage, about 36% of tax-6(p675) animals showed irregular and slowed DMP, while the remainder had regular DMP (Figure 2H), suggesting that tax-6(p675) is a weak allele. The fraction of the animals with irregular DMP appeared to increase with age, indicating that this phenotype might be age-dependent. This may also explain why tax-6(p675) animals were reported to have a normal defecation cycle in an earlier study (Lee et al., 2005).

Next, we investigated whether the reduced number of expulsion events was due to regular intervals with longer cycle lengths or if rhythmicity was entirely disrupted upon tax-6 knockdown. To assess this, we obtained ethograms of the DMP for N2 animals grown on control and tax-6 RNAi. While animals on control RNAi displayed regular cycles of pBoc, aBoc, and EMC, the tax-6 RNAi animals exhibited disrupted rhythmicity (Figure 3A and Figure 5). Most tax-6 knockdown animals lacked the pBoc and aBoc steps and had sporadic expulsion events. Isolated pBoc events were occasionally observed, indicating a complete loss of rhythmicity in tax-6 knockdown animals. Ethograms for tax-6(ok2065) animals also showed disrupted rhythmicity (Figure 3B and Figure 6). Although the number of expulsion events appeared higher in tax-6(ok2065) animals compared to tax-6 RNAi animals (Figure 5 and S6), these expulsion events seemed superficial, releasing little to no gut content. This suggested slow movement of gut content in tax-6(ok2065) animals, leading to constipation and intestinal bloating. We examined gut content movement by measuring the clearance of blue dye (erioglaucine disodium salt) from the gut. The clearance was significantly slower in tax-6(ok2065) animals compared to N2 animals (Figure 3C), indicating impaired gut content movement due to the loss of tax-6. Similarly, tax-6 knockdown animals also showed significantly slowed gut content movement (Figure 3D).

Calcineurin inhibition disrupts the C. elegans defecation motor program (DMP)

(A) Representative DMP ethograms of N2 animals after growth on the empty vector (EV) control and tax-6 RNAi bacteria till 1-day-old stage. Each dot represents a second, and each row represents a minute. ‘p’, ‘a’, and ‘x’ represent posterior body-wall muscle contraction, anterior body-wall muscle contraction, and expulsion muscle contraction, respectively.

(B) Representative DMP ethograms of N2 and tax-6(ok2065) animals after their growth on E. coli OP50 at 20°C till the 1-day-old adult stage.

(C) Plots of the change in the percent of animals with blue dye in their gut with time for 1-day-old adult N2 and tax-6(ok2065) animals (n = 3 biological replicates).

(D) Plots of the change in the percent of animals with blue dye in their gut with time for 1-day-old adult N2 animals grown on EV control and tax-6 RNAi bacteria (n = 3 biological replicates).

We then explored whether the disruption of DMP rhythmicity due to tax-6 knockdown affected P. aeruginosa responses similarly to longer but regular DMP cycles. To do this, we studied P. aeruginosa colonization in clk-1(qm30) and isp-1(qm150) mutants, which have regular but extended DMP cycles (Feng et al., 2001; Wong et al., 1995). Interestingly, both clk-1(qm30) and isp-1(qm150) mutants showed significantly reduced intestinal colonization by P. aeruginosa compared to N2 animals (Figure 7A-D). This reduced colonization could be attributed to their significantly decreased pharyngeal pumping rates (Wong et al., 1995; Yee et al., 2014), suggesting a lower intake of bacterial food in these mutants. While the survival of clk-1(qm30) animals on P. aeruginosa was comparable to N2 animals (Figure 7E), isp-1(qm150) animals exhibited significantly improved survival (Figure 7F). Conversely, knockdown of flr-1, nhx-2, and pbo-1 in N2 animals resulted in significantly reduced survival on P. aeruginosa compared to control RNAi (Figure 7G). Knockdown of these genes causes complete disruption of DMP rhythmicity, increasing gut colonization by P. aeruginosa (Singh and Aballay, 2019a). Overall, these findings demonstrated that calcineurin is crucial for maintaining the DMP ultradian clock, and its inhibition increases susceptibility to P. aeruginosa by disrupting the DMP.

Calcineurin inhibition enhances lifespan via DMP defects-induced calorie restriction

Defects in the C. elegans DMP lead to reduced nutrient absorption, resulting in decreased intestinal fat levels (Sheng et al., 2015). We investigated whether tax-6 knockdown similarly reduces fat levels. To test this, we performed oil-red-O (ORO) staining on tax-6 RNAi animals. The knockdown of tax-6 significantly reduced fat levels (Figure 4A, B). Next, we examined whether the reduced fat levels in tax-6 knockdown animals mimic calorie restriction-like phenotypes. Calorie restriction is known to upregulate the expression of the FoxA transcription factor PHA-4 (Panowski et al., 2007). Consistent with this, tax-6 knockdown also resulted in the upregulation of pha-4 expression levels (Figure 4C). Because calorie restriction enhances lifespan (Kaeberlein et al., 2006; Lakowski and Hekimi, 1998; Panowski et al., 2007), we explored whether calcineurin inhibition could similarly increase lifespan through calorie restriction mechanisms. The knockdown of tax-6 in the genetic model of calorie restriction, eat-2(ad465) (Lakowski and Hekimi, 1998), did not increase the lifespan (Figure 4D).

Calcineurin knockdown enhances lifespan via DMP defects-mediated calorie restriction

(A) Representative photomicrographs of oil-red-O (ORO) stained 1-day-old adult N2 animals grown on the empty vector (EV) control and tax-6 RNAi bacteria at 20°C. Scale bar = 200 µm.

(B) Quantification of ORO intensity per animal of 1-day-old adult N2 animals grown on the EV control and tax-6 RNAi bacteria at 20°C. The values are area normalized for each animal. ***p<0.001 via the t-test (n = 10 worms each).

(C) Quantitative reverse transcription-PCR (qRT-PCR) for pha-4 mRNA levels in N2 animals grown on the EV control and tax-6 RNAi bacteria at 20°C until 1-day-old adults. ***p < 0.001 (n = 3 biological replicates).

(D) Representative survival plots of eat-2(ad465) animals grown on the bacteria for RNAi against tax-6 along with the EV control at 20°C. Day 0 represents young adults. n.s., nonsignificant (n = 3 biological replicates; animals per condition per replicate > 80).

(E) Representative survival plots of N2, tax-6(ok2065), eat-2(ad465), and eat-2(ad465);tax-6(ok2065) animals grown on E. coli OP50 at 20°C. Day 0 represents young adults. p<0.001 for tax-6(ok2065), eat-2(ad465), and eat-2(ad465);tax-6(ok2065) compared to N2. p<0.05 for eat-2(ad465);tax-6(ok2065) compared to tax-6(ok2065) (n = 3 biological replicates; animals per condition per replicate > 65).

(F-J) Representative survival plots of aak-2(ok524) (E), raga-1(ok386) (F), rsks-1(ok1255) (G), daf-16(mu86) (H), and nhr-49(nr2041) (I) animals grown on the bacteria for RNAi against tax-6 along with the EV control at 20°C. Day 0 represents young adults. p<0.001 for all the plots (n = 3 biological replicates; animals per condition per replicate > 60).

Since eat-2(ad465) animals consume fewer bacteria, it is possible that their feeding-based RNAi efficiency is reduced, leading to a lack of responsiveness to tax-6 RNAi. To assess RNAi efficiency in eat-2(ad465) animals, we exposed them to act-5 and bli-3 RNAi, which are known to cause developmental arrest and a blistered cuticle, respectively (Gahlot and Singh, 2024). Similar to N2 animals, eat-2(ad465) mutants were sensitive to the toxic effects of act-5 and bli-3 knockdown (Figure 8), indicating that eat-2(ad465) animals are not defective in RNAi. To further confirm that calcineurin inhibition enhances lifespan via calorie restriction, we examined the lifespan of eat-2(ad465);tax-6(ok2065) double mutants. While eat-2(ad465) animals displayed increased lifespan compared to N2 controls, eat-2(ad465);tax-6(ok2065) double mutants had a lifespan similar to tax-6(ok2065) animals (Figure 4E). These results suggested that calcineurin inhibition promotes lifespan extension through calorie restriction mechanisms.

The mechanisms of lifespan enhancement by dietary or calorie restriction are complex and involve several different downstream pathways (Chamoli et al., 2014; Greer and Brunet, 2009; Hansen et al., 2008). Further, different pathways are operational under different dietary regimens (Greer and Brunet, 2009). Therefore, we investigated the involvement of pathways associated with calorie restriction-mediated lifespan extension in response to calcineurin inhibition. Specifically, we analyzed the interactions of the AMP-activated kinase, aak-2 (Greer and Brunet, 2009), the mTOR pathway, raga-1 (Robida-Stubbs et al., 2012; Zhang et al., 2019), ribosomal S6 kinase, rsks-1 (Chen et al., 2009; Selman et al., 2009; Zhang et al., 2019), the FOXO transcription factor, daf-16 (Greer and Brunet, 2009), and the nuclear hormone receptor, nhr-49 (Chamoli et al., 2014) with calcineurin inhibition in regulating lifespan. Mutants defective in aak-2 appeared to have only a partial increase in lifespan upon calcineurin inhibition (Figure 4F). In contrast, loss-of-function mutants of raga-1, rsks-1, daf-16, and nhr-49 exhibited enhanced lifespan upon calcineurin inhibition (Figure 4G-J).

Calcineurin knockdown enhances lifespan via HLH-30 and NHR-8

To further investigate longevity pathways regulated by calorie restriction that may enhance lifespan downstream of calcineurin inhibition, we examined the role of the TFEB ortholog HLH-30. HLH-30 is essential for the starvation response and triggers lipid depletion under nutrient-deprived conditions (O’Rourke and Ruvkun, 2013). Knockdown of tax-6 did not extend the lifespan of hlh-30(tm1978) animals (Figure 5A), indicating that HLH-30 is required for the increased lifespan observed with calcineurin inhibition. To determine whether HLH-30 is necessary for DMP defect-mediated lifespan extension, we assessed the lifespan of hlh-30(tm1978) animals upon knockdown of flr-1, nhx-2, and pbo-1. Knockdown of these genes did not alter the lifespan of hlh-30(tm1978) animals compared to control RNAi (Figure 5A), except for a slight increase in lifespan with nhx-2 knockdown. The hlh-30(tm1978) animals were found to be sensitive to RNAi (Figure 9), suggesting that the lack of lifespan change with tax-6, flr-1, and pbo-1 RNAi was not due to reduced RNAi efficiency in hlh-30(tm1978) animals. Additionally, we confirmed that tax-6 knockdown caused DMP defects and intestinal bloating in hlh-30(tm1978) animals (Figure 10A-C). We also explored whether HLH-30 affected fat depletion due to DMP defects. The ORO levels declined significantly in hlh-30(tm1978) animals upon the knockdown of tax-6 and other DMP-regulating genes (Figure 5B, C). The reduction in ORO levels was less pronounced in tax-6 knockdown animals compared to those with other DMP-regulating gene knockdowns.

Calcineurin inhibition enhances lifespan via HLH-30 and NHR-8

(A) Representative survival plots of hlh-30(tm1978) animals grown on bacteria for RNAi against tax-6, flr-1, pbo-1, and nhx-2 along with the empty vector (EV) control at 20°C. Day 0 represents young adults. p values for tax-6, flr-1, pbo-1, and nhx-2 compared to EV are <0.01, <0.001, nonsignificant, and <0.001, respectively (n = 3 biological replicates; animals per condition per replicate > 60).

(B) Representative photomicrographs of oil-red-O (ORO) stained 1-day-old adult hlh-30(tm1978) animals grown on the EV control, tax-6, flr-1, pbo-1, and nhx-2 RNAi bacteria at 20°C. Scale bar = 200 µm.

(C) Quantification of ORO intensity per animal of 1-day-old adult hlh-30(tm1978) animals grown on the EV control, tax-6, flr-1, pbo-1, and nhx-2 RNAi bacteria at 20°C. The values are area normalized for each animal. ***p<0.001 via the t-test (n = 20 worms each).

(D) Representative survival plots of nhr-8(ok186) animals grown on bacteria for RNAi against tax-6, flr-1, pbo-1, and nhx-2 along with the EV control at 20°C. Day 0 represents young adults. p values for tax-6, flr-1, pbo-1, and nhx-2 compared to EV are nonsignificant, <0.05, nonsignificant, and <0.05, respectively (n = 3 biological replicates; animals per condition per replicate > 70).

(E) Representative survival plots of daf-9(rh50) animals grown on the bacteria for RNAi against tax-6 along with the EV control at 20°C. Day 0 represents young adults. p<0.001 (n = 3 biological replicates; animals per condition per replicate > 80).

(F) Representative photomicrographs of ORO stained 1-day-old adult nhr-8(ok186) animals grown on the EV control, tax-6, flr-1, pbo-1, and nhx-2 RNAi bacteria at 20°C. Scale bar = 200 µm.

(G) Quantification of ORO intensity per animal of 1-day-old adult nhr-8(ok186) animals grown on the EV control, tax-6, flr-1, pbo-1, and nhx-2 RNAi bacteria at 20°C. The values are area normalized for each animal. ***p<0.001 via the t-test (n = 20 worms each).

(H) Quantitative reverse transcription-PCR (qRT-PCR) for nhr-8 mRNA levels in N2 animals grown on the EV control, tax-6, flr-1, pbo-1, and nhx-2 RNAi bacteria at 20°C until 1-day-old adults. ***p < 0.001 via the t-test (n = 3 biological replicates).

Steroid hormone signaling mediated by NHR-8 is known to regulate lifespan extension under calorie restriction (Thondamal et al., 2014). We tested whether NHR-8 was necessary for lifespan extension induced by calcineurin inhibition. Indeed, tax-6 knockdown did not extend the lifespan of nhr-8(ok186) animals (Figure 5D). While the specific ligands of NHR-8 are not well-defined (Magner et al., 2013), steroid hormones produced by the cytochrome P450 enzyme DAF-9 have been shown to enhance lifespan via NHR-8 under calorie restriction (Thondamal et al., 2014). Therefore, we investigated whether DAF-9 mediated the enhanced lifespan downstream of calcineurin inhibition. We found that tax-6 knockdown enhanced the lifespan of daf-9(rh50) animals (Figure 5E), indicating that DAF-9-produced steroid hormones were not required for lifespan increment by calcineurin inhibition. We also examined the lifespan of nhr-8(ok186) animals upon knockdown of flr-1, nhx-2, and pbo-1. Knockdown of these genes did not affect the lifespan of nhr-8(ok186) animals compared to control RNAi (Figure 5D). As with hlh-30(tm1978) animals, the lack of lifespan change in nhr-8(ok186) animals was not due to RNAi insensitivity, as they were responsive to RNAi (Figure 9). Furthermore, tax-6 knockdown resulted in DMP defects in nhr-8(ok186) animals (Figure 10D). We investigated whether NHR-8 influenced fat depletion upon DMP defects. A significant reduction in fat levels was observed in nhr-8(ok186) animals upon knockdown of tax-6, flr-1, nhx-2, and pbo-1 (Figure 5F, G), suggesting that NHR-8 functions downstream of or in parallel with fat depletion.

Since calorie restriction can enhance the expression of genes that promote lifespan, such as pha-4 (Panowski et al., 2007), we examined whether tax-6 knockdown modulates nhr-8 mRNA levels. We observed a significant increase in nhr-8 mRNA levels following tax-6 knockdown (Figure 5H). To determine whether this upregulation was due to intestinal bloating, we assessed nhr-8 mRNA levels in flr-1, nhx-2, and pbo-1 knockdown animals. Knockdown of these genes also increased nhr-8 mRNA levels (Figure 5H), indicating that intestinal bloating induces nhr-8 upregulation. Thus, calcineurin inhibition likely enhances lifespan via NHR-8 by increasing its expression.

Given that HLH-30 and NHR-8 are essential for lifespan extension upon calcineurin inhibition, we investigated whether these pathways also influence survival in response to P. aeruginosa infection following calcineurin knockdown. Both hlh-30(tm1978) and nhr-8(ok186) animals showed significantly reduced survival upon tax-6 RNAi (Figure 6A, B). These findings suggested that the reduced survival on P. aeruginosa following calcineurin inhibition is independent of HLH-30 and NHR-8 and is more likely due to increased gut colonization by P. aeruginosa resulting from DMP defects (Figure 6C).

Calcineurin inhibition-mediated effects on lifespan and survival on P. aeruginosa are mediated by distinct mechanisms

(A)-(B) Representative survival plots of hlh-30(tm1978) (A) and nhr-8(ok186) (B) animals on P. aeruginosa PA14 at 25°C after treatment with the empty vector (EV) control and tax-6 RNAi. p<0.001 for both the plots (n = 3 biological replicates; animals per condition per replicate > 85).

(C) Model depicting the mechanism of increased lifespan and enhanced susceptibility to pathogen upon inhibition of calcineurin via the defects in the defecation motor program (DMP).

Discussion

We discovered that calcineurin is required for the C. elegans DMP. Calcineurin activity is regulated by intracellular calcium levels. Increased amounts of calcium ions activate the phosphatase activity of calcineurin via the binding of the calcium-sensing protein calmodulin to calcineurin (Klee et al., 1998). The rhythmic DMP cycle in C. elegans is known to be regulated by rhythmic calcium waves (Santo et al., 1999; Teramoto and Iwasaki, 2006). The calcium waves are regulated by the endoplasmic reticulum calcium channel, ITR-1, and mutations in the itr-1 gene affect defecation by preventing cytoplasmic calcium release (Santo et al., 1999). It is likely that calcium waves regulate DMP via calcineurin activities. Indeed, calcineurin is expressed in the enteric muscles that are required for contractions for DMP (Lee et al., 2005). It is interesting to note that tax-6 gain-of-function mutants are also known to have DMP defects (Lee et al., 2005). Therefore, optimum calcineurin activity appears to be crucial for maintaining a rhythmic DMP.

Calcineurin inhibition has been shown to extend C. elegans lifespan via multiple mechanisms (Dong et al., 2007; Dwivedi et al., 2009; Mair et al., 2011; Tao et al., 2013). Activation of autophagy has been shown to be one of the mechanisms of extended lifespan upon calcineurin inhibition (Dwivedi et al., 2009). We showed that defects in the DMP caused by calcineurin inhibition reduce lipid levels and mimic calorie restriction. Calorie restriction is known to induce autophagy (Morselli et al., 2010), and autophagy is required for calorie restriction-mediated lifespan extension (Hansen et al., 2008; Jia and Levine, 2007). Therefore, it is likely that reduced lipid levels because of intestinal bloating result in the activation of autophagy upon calcineurin inhibition. Importantly, the TFEB ortholog, HLH-30, is required for lipolysis and autophagy under starvation conditions (Lapierre et al., 2013; O’Rourke and Ruvkun, 2013). HLH-30 is also required for lifespan enhancement via autophagy and multiple longevity pathways (Lapierre et al., 2013; O’Rourke and Ruvkun, 2013). We observed that HLH-30 is also required for the increase in lifespan upon calcineurin inhibition. Because autophagy may regulate lifespan via multiple longevity pathways (Hansen et al., 2008; Lapierre et al., 2013), this could explain why calcineurin inhibition enhances lifespan via multiple mechanisms.

Dietary restriction-mediated lifespan extensions are complex and context-dependent (Greer and Brunet, 2009). Multiple longevity pathways have been identified under different paradigms of dietary restriction (Chamoli et al., 2020, 2014; Chen et al., 2009; Greer and Brunet, 2009; Lapierre et al., 2013; Matai et al., 2019; Selman et al., 2009; Thondamal et al., 2014). We found that tax-6 knockdown enhanced lifespan independent of multiple dietary restriction pathways, including raga-1, rsks-1, daf-16, and nhr-49. An earlier study identified that tax-6 knockdown enhanced lifespan via NHR-49 (Burkewitz et al., 2015). We currently do not understand the reasons for this discrepancy. We identified that tax-6 knockdown resulted in the upregulation of nhr-8 mRNA levels, and nhr-8 was required for the increased lifespan in tax-6 knockdown animals. NHR-8 has been shown to mediate calorie restriction-dependent lifespan via the regulation of xenobiotic responses (Chamoli et al., 2014; Verma et al., 2018). It remains to be studied whether NHR-8 and HLH-30 work in the same or different pathways downstream of calcineurin inhibition (Figure 6C).

Recent studies in different organisms have shown that gut bloating has profound effects on food-seeking behaviors, immunity, and lifespan (Duvall et al., 2019; Filipowicz et al., 2021; Kumar et al., 2019; Min et al., 2021; Singh and Aballay, 2019a, 2019b, 2019c). Several C. elegans mutants with defects in the DMP and bloated intestinal lumens are known to have dampened nutrient absorption, leading to reduced lipid deposition in the gut, mimicking calorie restriction (Sheng et al., 2015). It is possible that some of the effects of intestinal bloating on the host physiology are mediated by calorie restriction. Indeed, the neuropeptide Y receptors, which control a diverse set of behaviors, including appetite, are activated by gut bloating (Singh and Aballay, 2019c). Calorie restriction is known to induce neuropeptide Y, which might trigger feeding behaviors (Aveleira et al., 2015; de Rijke et al., 2005; Ferreira-Marques et al., 2016). The complete characterization of the physiological changes downstream of gut bloating may provide broad insights into the effects of gut physiology on behaviors, immunity, and lifespan.

Materials and methods

Bacterial strains

The following bacterial strains were used: Escherichia coli OP50, E. coli HT115(DE3), Pseudomonas aeruginosa PA14, and P. aeruginosa PA14 expressing green fluorescent protein (P. aeruginosa PA14-GFP). The cultures for these bacteria were grown in Luria-Bertani (LB) broth at 37°C.

C. elegans strains and growth conditions

C. elegans hermaphrodites were maintained on nematode growth medium (NGM) plates seeded with E. coli OP50 at 20°C unless otherwise indicated. Bristol N2 hermaphrodites were used as the wild-type control unless otherwise indicated. The strains used in the study are provided in the Key Resources Table. Some of the strains were obtained from the Caenorhabditis Genetics Center (University of Minnesota, Minneapolis, MN). The eat-2(ad465);tax-6(ok2065) strain was generated using a standard genetic cross. The fer-1(b232) hermaphrodites were maintained on E. coli OP50 at 15°C and were grown at 25°C prior to P. aeruginosa killing assays and longevity assays to obtain animals with unfertilized oocytes.

RNA interference (RNAi)

RNAi was used to generate loss-of-function phenotypes by feeding nematodes E. coli strain HT115(DE3) expressing double-stranded RNA homologous to a target gene. RNAi was carried out as described previously (Gokul and Singh, 2022; Ravi et al., 2023). Briefly, E. coli with the appropriate vectors were grown in LB broth containing ampicillin (100 μg/mL) at 37°C overnight and plated onto NGM plates containing 100 μg/mL ampicillin and 3 mM isopropyl β-D-thiogalactoside (IPTG) (RNAi plates). RNAi-expressing bacteria were allowed to grow overnight at 37°C. Gravid adults were transferred to RNAi-expressing bacterial lawns and allowed to lay eggs for 2 hours. The gravid adults were removed, and the eggs were allowed to develop at 20°C to 1-day-old adults for subsequent assays. The RNAi clones were from the Ahringer RNAi library and were verified by sequencing.

RNAi efficiency test

The RNAi efficiency test was carried out as described earlier (Gahlot and Singh, 2024; Ghosh and Singh, 2024). Briefly, the wild-type N2, eat-2(ad456), nhr-8(ok186), and hlh-30(tm1978) worms were synchronized by egg laying and grown on empty vector control, act-5, and bli-3 RNAi plates at 20°C. The sid-1(qt9) worms were used as RNAi-defective controls. After 72 hours, the animals were monitored for development defects and blisters on the cuticle for act-5 and bli-3 RNAi, respectively.

C. elegans longevity assays

Lifespan assays were performed on RNAi plates containing E. coli HT115(DE3) with the empty vector control and tax-6 RNAi clone in the presence of 50 µg/mL of 5-fluorodeoxyuridine (FUdR). Animals were synchronized on RNAi plates without FUdR and incubated at 20°C. At the late L4 larval stage, the animals were transferred onto the corresponding RNAi plates containing 50 µg/mL of FUdR and incubated at 20°C. For fer-1(b232) lifespan assays, the animals were synchronized on RNAi plates without FUdR and incubated at 25°C. The fer-1(b232) lifespan assays were carried out at 25°C without FUdR. Animals were scored every other day as live, dead, or gone. Animals that failed to display touch-provoked movement were scored as dead. Animals that crawled off the plates were censored. Young adult animals were considered as day 0 for the lifespan analysis. Three independent experiments were performed.

C. elegans killing assays on P. aeruginosa PA14

Bacterial cultures were prepared by inoculating individual bacterial colonies of P. aeruginosa into 2 mL of LB and growing them for 10-12 hours on a shaker at 37°C. Bacterial lawns were prepared by spreading 20 µL of the culture on the entire surface of 3.5-cm-diameter modified NGM agar plates (0.35% instead of 0.25% peptone). The plates were incubated at 37°C for 12-16 hours and then cooled to room temperature for at least 30 minutes before seeding with synchronized 1-day-old adult animals. The killing assays were performed at 25°C, and live animals were transferred daily to fresh plates. Animals were scored at times indicated and were considered dead when they failed to respond to touch.

P. aeruginosa-GFP colonization assay

Bacterial cultures were prepared by inoculating individual bacterial colonies of P. aeruginosa-GFP into 2 mL of LB and growing them for 10-12 hours on a shaker at 37°C. Bacterial lawns were prepared by spreading 20 µL of the culture on the entire surface of 3.5-cm-diameter modified NGM agar plates (0.35% instead of 0.25% peptone) containing 50 μg/mL of kanamycin. The plates were incubated at 37°C for 12 hours and then cooled to room temperature for at least 30 minutes before seeding with 1-day-old adult gravid adults. The assays were performed at 25°C. After 6 hours of incubation, the animals were transferred from P. aeruginosa-GFP plates to fresh E. coli OP50 plates and visualized within 5 minutes under a fluorescence microscope.

Quantification of intestinal bacterial loads

The intestinal bacterial loads were quantified by measuring colony-forming units (CFU) as described earlier (Singh and Aballay, 2019b, 2019a) with appropriate modifications. Briefly, P. aeruginosa-GFP lawns were prepared as described above. After 6 hours of exposure, the animals were transferred from P. aeruginosa-GFP plates to the center of fresh E. coli OP50 plates thrice for 10 minutes each to eliminate bacteria stuck to their body. Afterward, ten animals/condition were transferred into 50 μL of PBS containing 0.01% triton X-100 and ground using glass beads. Serial dilutions of the lysates (101, 102, 103, 104) were seeded onto LB plates containing 50 μg/mL of kanamycin to select for P. aeruginosa-GFP cells and grown overnight at 37°C. Single colonies were counted the next day and represented as the number of bacterial cells or CFU per animal. At least three independent experiments were performed for each condition.

Fluorescence imaging

Fluorescence imaging was carried out as described previously (Gokul and Singh, 2022; Ravi et al., 2023). Briefly, the animals were anesthetized using an M9 salt solution containing 50 mM sodium azide and mounted onto 2% agarose pads. The animals were then visualized using a Nikon SMZ-1000 fluorescence stereomicroscope.

Quantification of intestinal lumen bloating

The 1-day-old adult animals grown on the empty vector control and tax-6 RNAi were anesthetized using an M9 salt solution containing 50 mM sodium azide, mounted onto 2% agarose pads, and imaged. The diameter of the intestinal lumen was measured using the Zeiss Zen Pro 2011 software. At least ten animals were used for each condition.

Measurement of defecation motor program (DMP) rate

Wild-type N2 and tax-6(p675) animals were synchronized and grown at 20°C on E. coli OP50 for 4 and 5 days, respectively, before measuring the DMP rate. For RNAi experiments, N2 worms were synchronized on EV and tax-6 RNAi plates and grown for 4 days at 20°C before measuring the DMP rate. The DMP cycle length was scored by assessing the time between expulsions (which are preceded by posterior and anterior body wall muscle contraction and the contraction of enteric muscles in a normal regular pattern) (Thomas, 1990). The number of expulsion events in 20 minutes was measured for each worm. The DMP rate was recorded for 5-6 worms/condition.

Ethogram for DMP

Wild type N2 and tax-6(ok2065) worms were synchronized by egg laying and grown at 20°C on E. coli OP50 for 4 and 5 days, respectively. For RNAi experiments, N2 worms were synchronized and grown on empty vector control and tax-6 RNAi at 20°C for 4 days. The assay was performed at 25°C. The defecation motor program consists of three steps-posterior body-wall muscle contraction (p), anterior body-wall muscle contraction (a), and expulsion muscle contraction (x). Animals were observed for the exact time of these three contractions on a dissecting scope for a continuous period of 20 minutes. A computer program was developed to plot the ethogram where we manually input the timings of “p”, “a”, and “x” contractions.

Blue dye assay to assess functional defecation rate

Wild type N2 and tax-6(ok2065) worms were synchronized by egg laying and grown at 20°C on E. coli OP50 for 4 and 5 days, respectively, before measuring the functional defecation rate. For RNAi experiments, N2 worms were synchronized and grown on empty vector control and tax-6 RNAi at 20°C for 4 days. The functional defecation rate was assessed using a blue dye, erioglaucine disodium salt (5% wt/v), as an indicator of gut content. The time required by worms to clear the blue dye from the gut was used to calculate a functional defecation rate. Briefly, 50 worms from each condition were incubated in a solution containing blue dye and E. coli OP50 culture in a 1:1 ratio for 2 hours. Then, the worms were transferred to an NGM plate seeded with E. coli OP50. About 25-27 worms from each condition were transferred to a fresh NGM plate seeded with E. coli OP50. The functional defecation rate was assessed by observing the clearance of blue dye from the intestine of worms every 5 minutes over a period of 60 minutes. Three independent experiments were performed for each condition.

Pharyngeal pumping assay

For the pharyngeal pumping assay, 1-day-old adult animals grown on the empty vector control and tax-6 RNAi were used. The number of contractions of the terminal bulb of the pharynx was counted for 30 seconds per worm. The pumping rates for 20 worms were recorded for each condition.

Lipid staining

The oil-red-O (ORO) staining was carried out as described earlier (Lynn et al., 2015) with appropriate modification. Briefly, wild-type N2, hlh-30(tm1978), and nhr-8(ok186) animals were synchronized on empty vector control and tax-6 RNAi bacteria and grown at 20°C for 4 days. About 300-400 synchronized gravid adult worms from experimental plates were washed three times with 1X PBS plus 0.01% triton X-100 (PBST). To permeabilize the cuticle, 600 µL of 40% isopropanol was added to 100 µL of worm pellet and was rocked for three minutes. The animals were spun down at 500 rpm for 30 seconds, and then 600 µL of supernatant was removed. Subsequently, 600 µL of ORO working stock was added to the worm pellet to stain the worms and incubated at room temperature for 1 hour on a shaker. ORO working stock was freshly prepared by diluting the stock 0.5% ORO in isopropanol to 60% in water and rocked at room temperature for 2 hours, followed by the removal of debris with a 0.22 µm filter. Afterward, worm samples were pelleted, 600 µL of supernatant was removed, 600 µL of PBST was added, and the samples were rocked for 30 minutes at room temperature. After that, the worms were mounted on a 2% agarose pad and imaged. At least ten worms/condition were imaged. Three independent biological replicates were performed. The ORO intensity was quantified using Image J software.

RNA isolation and quantitative reverse transcription-PCR (qRT-PCR)

Animals were synchronized by egg laying. Approximately 40 N2 gravid adult animals were transferred to 9-cm RNAi plates seeded with E. coli HT115 expressing the appropriate vectors and allowed to lay eggs for 4-5 hours. The gravid adults were then removed, and the eggs were allowed to develop at 20°C for 96 hours. Subsequently, the animals were collected, washed with M9 buffer, and frozen in TRIzol reagent (Life Technologies, Carlsbad, CA). Total RNA was extracted using the RNeasy Plus Universal Kit (Qiagen, Netherlands). Residual genomic DNA was removed using TURBO DNase (Life Technologies, Carlsbad, CA). A total of 6 μg of total RNA was reverse-transcribed with random primers using the High-Capacity cDNA Reverse Transcription Kit (Applied Biosystems, Foster City, CA).

qRT-PCR was conducted using the Applied Biosystems One-Step Real-time PCR protocol using SYBR Green fluorescence (Applied Biosystems) on an Applied Biosystems 7900HT real-time PCR machine in 96-well-plate format. Twenty-five-microliter reactions were analyzed as outlined by the manufacturer (Applied Biosystems). The relative fold-changes of the transcripts were calculated using the comparative CT(2-ΔΔCT) method and normalized to pan-actin (act-1, -3, -4) as described earlier (Singh and Aballay, 2019a, 2017). The cycle thresholds of the amplification were determined using StepOnePlus software (Applied Biosystems). All samples were run in triplicate. The primer sequences are provided in the Key Resources Table.

Quantification and statistical analysis

The statistical analysis was performed with Prism 8 (GraphPad). All error bars represent mean±standard deviation (SD). The unpaired, two-tailed, two-sample t-test was used when needed, and the data were judged to be statistically significant when p < 0.05. In the figures, asterisks (*) denote statistical significance as follows: *, p<0.05, **, p<0.01, ***, p<0.001, as compared with the appropriate controls. The Kaplan-Meier method was used to calculate the survival fractions, and statistical significance between survival curves was determined using the log-rank test. All experiments were performed in triplicate.